---
title: "Response Surfaces (Intro)"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [5.5](), [Rmarkdown]()_

```{r}
library(dplyr)
library(ggplot2)
library(readr)
theme_set(theme_bw())
```

We’ve really pushed a particular recipe for all our hypothesis testing
approaches,

* Write down a model with various factor level effects

* Write down a sum-of-squares identity, and get the corresponding
degrees-of-freedom

* Get a test statistic and its reference distribution, for testing each factor

We’ve used this in ANOVA, RCBD, general factorial designs…

It’s been a while coming, but we’re finally going to have a new idea. The new
idea is to fit response surfaces. The idea is simple: use a flexible (nonlinear)
function from experimental inputs (combinations of factor levels) to the
response of interest.

* If we can assume the response varies smoothly as factor inputs are perturbed,
then this will work

* We’ll be able to use the fit to determine important influences as well as find
configurations that optimize the response (e.g., maximize profit[^Or if you are
disgruntled, minimize profit]).

* If we can assume that small changes in the levels of a factor don’t result in
very dramatic changes in the response, then we’ll have a good representation of
how varying factors affects the response.

How should we fit these flexible functions?

* Polynomial regression: include terms like $x_i^2, x_i^3, x_i^2 x_j, ...$

* Spline regression: Include polynomial terms, but split across different
regions of the input space. This is generally more stable than polynomial
regression.

* Really, you can use whatever function fitter that you want.

Unfortunately, this idea will have to wait till near the end of the course for a
more complete elaboration.

## Data Example

Let's look at this idea using the battery data from before. We'll treat
temperature as a continuous variable, so that it makes sense to talk about a
response surface[^really, a curve over temperature].

```{r}
battery <- read_table2("https://uwmadison.box.com/shared/static/vmxs2wcsdxkdjujp85nw5kvk83xz4gl9.txt") %>%
  mutate( Material = as.factor(Material) )
```

We fit a quadratic regression to define the surface. The result let's us make
predictions for battery life at temperatures that we haven't observed.

```{r}
fit <- lm(Life ~ Material * poly(Temperature, 2), data = battery)
surface <- expand.grid(
  Material = unique(battery$Material), 
  Temperature = seq(15, 125, by = 1)
  )
surface$Life <- predict(fit, surface)
ggplot() +
  geom_point(data = battery, aes(x = Temperature, y = Life)) +
  geom_line(data = surface, aes(x = Temperature, y = Life)) +
  facet_wrap(~ Material)
```

Compare the associated fit with Table 5.15.

```{r}
summary(fit)
```