---
title: "Common Distributions"
description: |
  Distributions that appear across experimental design.
author:
  - name: Kris Sankaran
    url: {}
date: 09-15-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [2.3](), [Rmarkdown](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/2021-08-03-week2-2/week2-2.Rmd)_

1. Don’t try to memorize the formulas for all the probability distributions! Instead, it's much more useful to learn,
* The relationships between distributions
* The basic shapes of the distributions (unimodal? nonnegative? ...)
* How their shapes change then their parameters are changed
We'll give a refresher of some common probability distributions in these notes.

2. **Chi-Square Distribution**. This distribution arises as the sum of squares
of standard normals. That is, if $z_[k] \sim \mathcal{N}\left(0, 1\right)$, then
$\sum_{k} z_{k}^2 \sim \chi^2_{K}$, a chi-square distribution with $K$-degrees
of freedom (d.f.).

3. This distributions claim to fame is that if $y_i \sim \mathcal{N}\left(\mu,
\sigma^2\right)$ independently, then

$$
\frac{1}{\sigma^2}\sum_{i = 1}^{n} \left(y_i- \bar{y}\right)^2 \sim \chi^2_{n -1}
$$
which is a nontrivial but very useful fact, since the expression on the right is
similar to the usual estimator for the sample standard deviation. We'll make use
of connection when we construct some common hypothesis tests.

4. **$t$ distribution**.  A $t$ distribution with $k$ d.f. arises as a ratio
between a normal and the square root of a chi-square with K d.f.,

$$
\frac{\mathcal{N}\left(0, 1\right)}{\sqrt{\frac{\chi^2_{K}}{K}}}
$$

This seems like an esoteric fact, but notice that the usual way of standardizing
the mean (when the true variance is unknown) has this form,

$$
\frac{\sqrt{n}\left(\bar{y} - \mu\right)}{S}
$$


5. **$F$ Distribution**. The $F$ distribution occurs as the ratio of independent
chi-squares (suitably rescaled), 

$$
F_{u, v} = \frac{\frac{1}{u}\chi^2_u}{\frac{1}{v}\chi^2_v}
$$

Since chi-squares arise whenever we have sums of squares, this distribution will
come in handy whenever we need to compare two different sums of squares.
	
```{r, preview = TRUE}
library(EBImage)
display(readImage("https://uwmadison.box.com/shared/static/dv5tvok0m9vkqqmkd3c0woam5is7gzse.png"))
```
