---
title: "$2^2$ Factorial Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r, echo = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
theme_set(theme_bw())
```

```{r, echo = FALSE, fig.margin = TRUE}
library(EBImage)
display(readImage("https://uwmadison.box.com/shared/static/e6sprneaeamlav35csswq68t62p0xney.png"))
display(readImage("https://uwmadison.box.com/shared/static/jst0ib619vjw386gnl2hs8pee5rkkbx1.png"))
```
1. A special case of general factorial design is the $2^K$ design. This arises
when there are $K$ factors, but only 2 levels for each factor. We assume $n$
samples at each configuration of factor levels. With only two levels for each
factor, these experiments aren’t useful for teasing out subtle variations across
levels of a factor. However they *are* useful for determining which of a large
number of factors might be worth investigating further (factor screening).

2. The simplest case of a $2^K$ design is when the number of factors $K = 2$. The
experimental design can be represented by a square.

```{r, echo = FALSE}
yield <- read_table2("https://uwmadison.box.com/shared/static/bfwd6us8xsii4uelzftg1azu2f7z77mk.txt") %>%
  mutate(
    A = as.factor(A),
    B = as.factor(B)
  )
```


### Notation

3. It will be handy to define ways of indexing corners of the square. One
approach is to write + or - for whether we are at a low or high level for that
factor. Alternatively, we can represent the corner by all the letters that are
at high levels.

```{r, echo = FALSE, fig.margin = TRUE}
ggplot(yield) +
  geom_point(
    aes(x = A, y = B, col = Yield),
    position = position_jitter(w = 0.1, h = 0.1)
  ) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom")
ggplot(yield) +
  geom_point(aes(x = A, y = Yield)) +
  facet_wrap(~B)
interaction.plot(yield$A, yield$B, yield$Yield)
```

| A | B | label |
|---|---|---------|
| - | - | (1)     |
| + | - | a       |
| - | + | b       |
| + | + | ab      |

4. We abuse notation and write $a$ to represent the *total* of the response
values at the corner +-, rather than just the index of that corner.

### Estimating effects

5. Since there are only two levels for each factor, there are transparent
formulas for estimating main and interaction effects.
  
    * The main effect for A summaries the average change in the response when A is
    activated. It is defined as
    \begin{align}
    A &= \frac{1}{2n}\left(\left(ab +a\right) - \left(b + (1)\right)\right)
    \end{align}
    which is the average of the responses on the edge where A is active minus the
    average when A is inactive. The definition for B is analogous.
    
```{r, echo = FALSE, fig.margin = TRUE}
display(readImage("https://uwmadison.box.com/shared/static/32pelq6o9jz7nnnavkdx1kjuxfjkvzrc.png"))
display(readImage("https://uwmadison.box.com/shared/static/hvedijshgaw7so91kqcl1l8eunml4m9v.png"))
```

    * The interaction effect measures the degree to which the effect of A changes
    depending on whether or not B is active. It is defined as
    \begin{align}
    AB &= \frac{1}{2n}\left[\left(ab - b\right) + \left(a - \left(1\right)\right)\right]
    \end{align}
    Notice that the role of A and B is symmetric — you could read the interaction as
    how the effect of B changes depending on whether A is active.
    
6. All these effect estimates can be summarized by our tabular notation,

|label | effect A | effect B | effect AB |
| --- |---|---|---------|
| (1) | - | - |  + |
| a | + | - | - |
| b | - | + | - |
| ab | + | + | + |
