---
title: "$2^2$ Factorial Designs"
description: |
  Two factors each with two levels.
author:
  - name: Kris Sankaran
    url: {}
date: 10-26-2021
output:
  distill::distill_article:
    self_contained: false
---

_Readings [6.1](https://www.wiley.com/en-us/Design+and+Analysis+of+Experiments%2C+10th+Edition-p-9781119492443#content-section), [Rmarkdown](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/2021-08-16-week8-1/week8-1.Rmd)_

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r, echo = FALSE}
library(ggplot2)
theme424 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(theme424)
```

```{r, echo = FALSE, fig.margin = TRUE}
library(EBImage)
display(readImage("https://uwmadison.box.com/shared/static/e6sprneaeamlav35csswq68t62p0xney.png"))
display(readImage("https://uwmadison.box.com/shared/static/jst0ib619vjw386gnl2hs8pee5rkkbx1.png"))
```
1. A special case of general factorial design is the $2^K$ design. This arises
when there are $K$ factors, but only 2 levels for each factor. We assume $n$
samples at each configuration of factor levels. With only two levels for each
factor, these experiments aren’t useful for teasing out subtle variations across
levels of a factor. However they *are* useful for determining which of a large
number of factors might be worth investigating further (factor screening).

2. The simplest case of a $2^K$ design is when the number of factors $K = 2$. The
experimental design can be represented by a square.

```{r}
library(readr)
library(dplyr)
yield <- read_table2("https://uwmadison.box.com/shared/static/bfwd6us8xsii4uelzftg1azu2f7z77mk.txt") %>%
  mutate_at(vars(-Yield), as.factor)
```


### Notation

3. It will be handy to define ways of indexing corners of the square. One
approach is to write + or - for whether we are at a low or high level for that
factor. Alternatively, we can represent the corner by all the letters that are
at high levels.

```{r}
ggplot(yield) +
  geom_point(aes(x = A, y = Yield)) +
  facet_wrap(~B)
interaction.plot(yield$A, yield$B, yield$Yield)
```

| A | B | label |
|---|---|---------|
| - | - | (1)     |
| + | - | a       |
| - | + | b       |
| + | + | ab      |

4. We abuse notation and write $a$ to represent the *total* of the response
values at the corner +-, rather than just the index of that corner.

### Estimating effects

5. Since there are only two levels for each factor, there are transparent
formulas for estimating main and interaction effects.
  
    * The main effect for A summaries the average change in the response when A is
    activated. It is defined as
    \begin{align}
    A &= \frac{1}{2n}\left(\left(ab +a\right) - \left(b + (1)\right)\right)
    \end{align}
    which is the average of the responses on the edge where A is active minus the
    average when A is inactive. The definition for B is analogous.
    
```{r, echo = FALSE}
display(readImage("https://uwmadison.box.com/shared/static/32pelq6o9jz7nnnavkdx1kjuxfjkvzrc.png"))
display(readImage("https://uwmadison.box.com/shared/static/hvedijshgaw7so91kqcl1l8eunml4m9v.png"))
```

    * The interaction effect measures the degree to which the effect of A changes
    depending on whether or not B is active. It is defined as
    \begin{align}
    AB &= \frac{1}{2n}\left[\left(ab - b\right) + \left(a - \left(1\right)\right)\right]
    \end{align}
    Notice that the role of A and B is symmetric — you could read the interaction as
    how the effect of B changes depending on whether A is active.
    
6. All these effect estimates can be summarized by our tabular notation,

|label | effect A | effect B | effect AB |
| --- |---|---|---------|
| (1) | - | - |  + |
| a | + | - | - |
| b | - | + | - |
| ab | + | + | + |
