---
title: "RCBD with Random Block Effects"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-06-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [4.1](), [Rmarkdown]()_

```{r}
include_graphics("https://uwmadison.box.com/s/gu58jurmhyraebsdyh4txlpyxegxn9xf")
```

As in standard random effects, sometimes the blocks are from a larger population
We care about a “resin effect,” but don’t care about each individual batch The
model is setup as before,



except now both jN(0, 2) and ijN(0, 2), all independently.

Observations are correlated within each block are correlated

Testing

We’re interested in whether any of the treatments have an effect,



We won’t show it, but it turns that

	

so we should reject the null when MSTreatment is much larger than MSE. In fact, as in the fixed block case,



so we can use the same F-distribution cutoff when testing whether any treatment effects are nonzero.

Estimation

As in random effects for the completely randomized design, we can use either the method of moments or maximum likelihood
The method of moments estimators are
	
	
Finding confidence intervals continues to be a challenge for the method of moments approach. If you want that, you should use maximum likelihood, as shown in the computer example accompanying these notes.
