---
title: "RCBD with Random Block Effects"
description: |
  The random effects analog of RCBD designs
author:
  - name: Kris Sankaran
    url: {}
date: 10-07-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = FALSE)
```

_Readings [4.1](https://www.wiley.com/en-us/Design+and+Analysis+of+Experiments%2C+10th+Edition-p-9781119492443#content-section), [Rmarkdown](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/2021-08-06-week5-3/week5-3.Rmd)_

1. As in standard random effects, sometimes the blocks are from a larger population. For example, in the medical device example, we care about a ``resin effect,'' but don’t care about each individual batch. 

2. The model is setup as before,

$$
y_{ij} = \mu + \tau_i + \beta_j + \epsilon_{ij}
$$
  except now both $\beta_j \sim \mathcal{N}\left(0, \sigma_{\beta}^2\right)$ and
  $\epsilon_{ij} \sim \mathcal{N}\left(0, \sigma^2\right)$, all independently.

```{r}
library(EBImage)
display(readImage("https://uwmadison.box.com/shared/static/gu58jurmhyraebsdyh4txlpyxegxn9xf.png"))
```

3. Observations are correlated within each block are correlated. The calculation
to demonstrate this is similar to the one used to show that observations $i$
within levels of an ordinary random effects model are correlated.

### Hypothesis Testing

4. We’re interested in whether any of the treatments have an effect,

\begin{align*}
H_0 &: \tau_1 = \dots = \tau_a = 0 \\
H_{1} &: \tau_{i} \neq 0 \text{ for at least one } i
\end{align*}

5. We won’t show it, but it turns that
\begin{align*}
\mathbf{E}\left[MS_{\text{Treatment}}\right] &= \sigma^2 + \frac{b \sum_{i = 1}^{a} \tau_i^2}{a - 1} \\
\mathbf{E}\left[MS_{\text{Block}}\right] &= \sigma^2 + a \sigma^2_{\beta} \\
\mathbf{E}\left[MS_{E}\right] &= \sigma^2
\end{align*}
so we should reject the null when $MS_{\text{Treatment}}$ is much larger than
$MS_E$. 

6. In fact, as in the fixed block case,
$$
\frac{MS_{\text{Treatment}}}{MS_{E}} \sim F\left(a - 1, \left(a - 1\right)\left(b - 1\right)\right)
$$
  so we can use the same F-distribution cutoff when testing whether any treatment effects are nonzero.

### Estimation

7. As in random effects for the completely randomized design, we can use either
the method of moments or maximum likelihood. The method of moments estimators are
\begin{align*}
\hat{\sigma}	&= MS_{E} \\
\hat{\sigma}^2_{\beta} &= \frac{1}{a}\left[MS_{\text{Block}} - MS_{E}\right]
\end{align*}
	
8. Finding confidence intervals continues to be a challenge for the method of
moments approach. In this case, maximum likelihood is preferred. This method is
shown in the computer example accompanying these notes.

```{r}
opts_chunk$set(echo = TRUE)
```

### Code Example

9. We'll continue with the graft dataset. This reads in the data as in the
previous notes.

```{r}
library(readr)
library(tidyr)
graft <- read_table2("https://uwmadison.box.com/shared/static/0ciblk4z2f3k6zizbj4plg3q33w1d0n6.txt") %>%
  pivot_longer(-Pressure, names_to = "batch", values_to = "yield")
graft$Pressure <- as.factor(graft$Pressure)
```

10. As with ordinary random effects ANOVA, we can fit an RCBD with random block
effects using `lmer`. Notice that we only use the `(1 | -)` notation for the
batch variable. This means that the batches are viewed as random but pressure is
considered fixed. This is why in the summary output, we have a variance
component $\sigma^2_{\beta}$ for the batches, but separate fixed effects for
across pressure levels.

```{r}
library(lme4)
fit <- lmer(yield ~ Pressure + (1 | batch), data = graft)
fit
```
