---
title: "$2^{K - p}$ Fractional Factorial Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-17-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```


_Readings [8.3, 8.4](), [Rmarkdown]()_


```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(reshape2)
library(EBImage)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r helper_funs}
code <- function(x) ifelse(x == '-', -1, 1)
plot_aliases <- function(fit, n_keep = NULL, n_sample=16) {
  X <- model.matrix(fit)
  aliases <- (1 / n_sample) * t(X) %*% X
  
  if (!is.null(n_keep)) {
    aliases <- aliases[1:n_keep, 1:n_keep]
  }
  
  maliases <- melt(aliases)
  ggplot(maliases) +
    geom_tile(aes(x = Var1, y = Var2, fill = as.factor(value)), col = "black") +
    scale_fill_brewer(palette = "Set3") +
    theme(
      legend.position = "none",
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_text(size = 6),
      axis.text.x = element_text(angle = -90)
    ) +
    coord_fixed()
}
daniel_plot <- function(effects, probs = c(0.4, 0.6)) { 
  qq <- qqnorm(effects, datax = TRUE)
  qqline(effects, col = "red", probs = probs, datax = TRUE)
  text(qq$x, qq$y, names(effects), pos=1)
}
```

$2^{K - 1}$ designs allow us to draw conclusions about $K$ factors using only
half of the runs that would be required for a full $2^{K}$ factorial. It’s
possible to generalize these ideas so that smaller fractions ($\frac{1}{4},
\frac{1}{8}, \dots$ of samples are required.  When only a fraction
$\frac{1}{2^{p}}$ is required, we call the resulting design a $2^{K - p}$
fractional factorial design.

### Recipe

* We can begin constructing a $2^{K - p}$ design by building a full factorial
design from $K - p$ of the factors

	* The associated full factorial is called the _basic design_
	
* We then need $p$ generating relations, which confound the remaining $p$
factors with terms from the full factorial design

	* The _complete defining relation_ for a design is the set of columns that equal $I$, the identity column
	
	* These can be found by looking at generating relations and their products
	
Here’s the recipe in action.

* Suppose we want a $2^{6 - 2}$ design.

	* Study 6 factors using 16 (not 64) runs.
	
* Create a full factorial on the first four factors

	* I’m only writing contrasts for the main effects
	
	* You can get the corresponding corner of the cube by isolating the positive
	signs for that row. E., the first two rows are $(1)$ and $a$, and the last two
	are $bcd$ and $abcd$.
	
| A | B | C	| D |
| --- | --- | --- | --- |
| - | - | - | - |
| + | - | - | - |
| - | + | - | - |
| - | - | + | - |
| - | - | - | + |
| + | + | - | - |
| + | - | + | - |
| + | - | - | + |
| - | + | + | - |
| - | + | - | + |
| - | - | + | + |
| + | + | + | - |
| + | + | - | + |
| + | - | + | + |
| - | + | + | + |
| + | + | + | + |

* Study the last two factors through the defining relations $E = ABC$ and $F =
BCD$. Our hope is that by defining them as high-order interactions, we have a
chance at higher resolution.

	* The associated complete defining relations are $I = ABCE = BCDF = ADEF$.
	
	* To see this, notice $E^2 = ABCE$, but any term squared is just $I$. The last
	relation comes from multiplying the two previous ones together.
	
The resulting design is

| A | B | C	| D | E = ABC | F = BCD |
| --- | --- | --- | --- | --- | --- |
| - | - | - | - | - | - |
| + | - | - | - | + | - |
| - | + | - | - | + | + |
| - | - | + | - | + | + |
| - | - | - | + | - | + |
| + | + | - | - | - | + |
| + | - | + | - | - | + |
| + | - | - | + | + |  + |
| - | + | + | - | - | - |
| - | + | - | + | + | - |
| - | - | + | + | + | - |
| + | + | + | - | + | - |
| + | + | - | + | - | - |
| + | - | + | + | - | - |
| - | + | + | + | - | + |
| + | + | + | + | + | + |

Let’s analyze this design,

* The alias groups are complicated looking, but they can be found by mindlessly
multiplying the defining relations by each of the factors and combinations of
factors.

	* In practice, you would use code to find the aliases (see below)
	
\begin{align*}
I&=A B C E=B C D F=A D E F \\
A&=B C E=D E F=A B C D F \\
B&=A C E=C D F=A B D E F \\
C&=A B E=B D F=A C D E F \\
D&=B C F=A E F=A B C D E \\
E&=A B C=A D F=B C D E F \\
F&=B C D=A D E=A B C E F \\
A B&=C E=A C D F=B D E F \\
A C&=B E=A B D F=C D E F \\
A D&=E F=B C D E=A B C F \\
A E&=B C=D F=A B C D E F \\
A F&=D E=B C E F=A B C D \\
B D&=C F=A C D E=A B E F \\
B F&=C D=A C E F=A B D E \\
A C D&=B D E=A B F=C E F \\
A B D&=C D E=A C F=B E F
\end{align*}

* From the alias groups, we can tell that the resolution is 4. Two-way
interactions are confounded with one another, but not with any main effects.

For an exercise, you can try going through this process using an alternative
confounding structure: Set $E = ABCD$, and $F = ABC$. It’s somewhat tedious, but
will build your confidence with this type of design.

### Code Example

Let’s use the $2^{6 - 2}$ design that we just constructed on a dataset about
injection molding. `r tufte::margin_note("Apparently this is how paper clips are made.") 
In any case, it’s the example in the book (Example 8.4). The 6 factors are,
* A: mold temperature
* B: screw speed
* C: holding time
* D: cycle time
* E: gate size
* F: holding pressure
```{r}
injection <- read_table2("https://uwmadison.box.com/shared/static/uxd6sryqz32gbubwfhbdvsnqqkplqqef.txt") %>%
  mutate_at(vars(-Shrinkage), code)
head(injection)
```

```{r, fig.cap="Visualizing a $2 ^ {6 - 2}$ design for the injection experiment. Each column is a factor, each row is a run."}
mdesign <- injection %>%
  mutate(id = row_number()) %>%
  select(-Shrinkage) %>%
  melt(id.vars = "id")
ggplot(mdesign) +
  geom_tile(aes(x = variable, y = id, fill = as.factor(value))) +
  coord_fixed() +
  scale_fill_brewer(palette = "Set3")
```

We can look at the aliasing structure by inspecting the design matrix.

```{r, fig.margin = FALSE, fig.fullwidth = TRUE, fig.caption = "Aliases in the $2^{6 - 2}$ Design."}
fit <- lm(Shrinkage ~ A * B * C * D * E * F, data = injection)
plot_aliases(fit)
```


```{r, fig.margin = FALSE, fig.caption = "A zoomed-in version of the aliases from before, focusing only on main effects and second-order interactions."}
plot_aliases(fit, 22)
```
Let’s make a Daniel plot and look at effects.

```{r}
daniel_plot(2 * coef(fit)[-1])
```

Let's refit to just the submodel.

```{r}
fit <- lm(Shrinkage ~ A * B, data = injection)
anova(fit)
injection <- injection %>%
  mutate(residual = resid(fit))
ggplot(injection) +
  geom_point(aes(x = C, y = residual, col = Shrinkage)) +
  scale_color_viridis_c() +
  facet_grid(A ~ B)
```

Notice that the residuals have lower variance at the setting $C = -1$. Even
though $C$ doesn’t have a strong effect, that might be a useful fact, in case
you want to select a configuration with the optimal response and low variability
around that response.

### Evaluating Designs

In general, there will be several ways of constructing any $2^{p}$ design. How
do we know which one to use? Here are some criteria.

* Resolution: We’ve discussed this before. Remember, a design has resolution $R$
if no $p$-factor effect is aliased with an effect with fewer than $R - p$
factors.

* Aberration: This is the number of defining words of length equal to the resolution. Lower is better.

	* The example above had words $ABCE, BCDF$ and $ADEF$ and resolution 4, so aberration is 3.