---
title: "Optimizing Multiple Responses"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 12-02-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [11.3](), [Rmarkdown]()_



```{r}
library(dplyr)
library(EBImage)
library(desirability)
library(ggplot2)
library(readr)
library(reshape2)
library(rsm)
library(stringr)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r, fig.margin = FALSE}
chem <- read_csv("https://uwmadison.box.com/shared/static/nbaj1m8j7tuaqmznjlrsgbzyhp9k61i8.csv")
std <- function(x) {
  (x - mean(x)) / sd(x)
}
mchem <- chem %>%
  mutate_at(vars(yield, viscosity, molecular_weight), std) %>%
  melt(measure.vars = c("yield", "viscosity", "molecular_weight"))
```

```{r}
chem_coded <- coded.data(chem, time_coded ~ (time - 35) / 5, temp_coded ~ (temp - 155) / 5)
fits <- list(
  "yield" = rsm(yield ~ SO(temp_coded, time_coded), data = chem_coded),
  "viscosity" = rsm(viscosity ~ SO(temp_coded, time_coded), data = chem_coded),
  "molecular_weight" = rsm(molecular_weight ~ SO(temp_coded, time_coded), data = chem_coded)
)
```
```{r}
contour(fits[[1]], ~ time_coded + temp_coded, image = TRUE, asp = 1)
```
```{r}
contour(fits[[2]], ~ time_coded + temp_coded, image = TRUE, asp = 1)
```
```{r, fig.cap = "Three separate response surfaces, fit to yield, viscosity, and molecular weight, respectively."}
contour(fits[[3]], ~ time_coded + temp_coded, image = TRUE, asp = 1)
```

In more complicated systems, we may want to optimize several objectives
simultaneously. More often than not, the goals will be at odds with one another.
For example, in our running chemical process example, we want to maximize yield
while maintaining a target viscosity and minimizing molecular weight. How can we
use response surface methods when we have several competing objectives?

\vspace{5mm}

```{r, fig.margin = FALSE}
ggplot(mchem) +
  geom_point(
    aes(x = time, y = temp, col = value),
    position = position_jitter(w = 0.3, h = 0.3)
  ) +
  facet_grid(. ~ variable) +
  coord_fixed() +
  scale_color_viridis_c() +
  theme(legend.position = "none")
```

### Overlaying Contours

* The most direct approach is to simply fit several response surfaces.

* Visually inspect results to find factor configurations with desirable values across each response.

### Constrained Optimization

Whenever visual inspection is challenging, mathematical formalizations can offer
support. One idea is to frame the multiple response surface problem as a
constrained optimization.

* Define acceptable ranges for responses  $y_{2}\left(x\right), \dots,
y_{R}\left(x\right)$ that are important, but not our main focus.

* Optimize the response $y_{1}\left(x\right)$ that’s our main focus.

Formally, we look for a configuration of factors $x_{\ast}$ that solves the
optimization
\begin{align*}
\underset{x}{\text{maximize}}\medspace &y_{1}\left(x\right) \\
\text{subject to }\medspace &\left(y_{2}\left(x\right), \dots, y_{R}\left(x\right)\right) \in \mathcal{C}
\end{align*}

where $C$ is the predefined acceptable region for the secondary responses.

### Desirability Functions

The main downside of the constrained optimization approach is that it forces us
to choose one response to prioritize over all others. What if we care about each
response more or less equally?

One idea is to optimize a sort of (geometric) averaged response,
\begin{align*}
\underset{x}{\text{maximize}}\medspace \left[\prod_{r = 1}^{R} y_{r}\left(x\right)\right]^{\frac{1}{R}}
\end{align*}

```{r, fig.cap = "Example desirability functions, for maximizing, minimizing, and achieving a target response."}
x_grid <- seq(-1, 1, .01)
example_funs <- data.frame(
  x = x_grid,
  d_max_0.5 = predict(dMax(0, 1, 0.5),  x_grid),
  d_max_1 = predict(dMax(0, 1, 1),  x_grid),
  d_max_2 = predict(dMax(0, 1, 2), x_grid),
  d_min_0.5 = predict(dMin(-1, 0, 0.5), x_grid),
  d_min_1 = predict(dMin(-1, 0, 1), x_grid),
  d_min_2 = predict(dMin(-1, 0, 2), x_grid),
  d_target_0.5 = predict(dTarget(-1, 0, 1, 0.5, 0.5), x_grid),
  d_target_1 = predict(dTarget(-1, 0, 1), x_grid),
  d_target_2 = predict(dTarget(-1, 0, 1, 2, 2), x_grid)
) %>%
  melt(id.vars = c("x")) %>%
  mutate(
    type = str_replace(variable, "[\\.0-9]+", ""),
    scale = str_extract(variable, "[\\.0-9]+")
  )
ggplot(example_funs) +
  geom_line(aes(x = x, y = value, col = scale)) +
  scale_color_brewer(palette = "Set2") +
  facet_grid(type ~ .)
```



The issue with this idea is that it treats all responses exactly equally. What
if we want to maximize some, but minimize others? What if we want some to be
near some target value?

The solution is to use _desirability functions_. A few are plotted below. You
can adjust their shape so that the $r^{th}$ desirability function is large for
the values of $y_{r}\left(x\right)$ which are good (sloping down when you want
to minimize, sloping up when you want to maximize).

Then, instead of optimizing the raw averaged response, we optimize the averaged
response after first passing through the desirability functions,
\begin{align*}
\underset{x}{\text{maximize}} \medspace \left[\prod_{r = 1}^{R} d_{r}\left(y_{r}\left(x\right)\right)\right]^{\frac{1}{R}}
\end{align*}

### Data Example

For the chemical problem, we can define a desirability function per response,
and then combine them into an overall objective. This is done using the
`desirability` package.

```{r, echo = TRUE}
d_yield <- dMax(70, 95) # min / max / scale
d_viscosity <- dTarget(55, 65, 75) # min / target / max
d_weight <- dMin(2750, 4000) # min / max
objective <- dOverall(d_yield, d_viscosity, d_weight)
```

Now, we can apply this objective function to predictions made by the three
response surface fits we made above, in the section on overlaying contours.
We're evaluating the predictions over a grid of values in the range of the coded
time and temperature.

```{r}
coded <- as.data.frame(chem_coded)
x_grid <- expand.grid(
  time_coded = seq(min(coded$time_coded), max(coded$time_coded), .1),
  temp_coded = seq(min(coded$temp_coded), max(coded$temp_coded), 0.1)
)
```

```{r, echo = TRUE}
desirabilities <- cbind(
  x_grid,
  y1 = predict(fits[[1]], x_grid),
  y2 = predict(fits[[2]], x_grid),
  y3 = predict(fits[[3]], x_grid)
)
y_hat <- desirabilities %>% 
  select(starts_with("y"))
desirabilities$objective <- predict(objective, y_hat)
```

At this point, we can plot the desirability of each point in the time /
temperature configuration space. Note that a second mode emerges along the
low-temperature region -- these are regions which have good viscosity and
molecular weight properties, even though their yield isn't as high. Try
increasing the scale for yield to see what happens when you make the requirement
for high yield more stringent.

```{r, fig.margin = FALSE, fig.cap = "Overall desirability, considering yield, viscosity, and molecular weight."}
ggplot(desirabilities) +
  geom_tile(
    aes(x = time_coded, y= temp_coded, fill = objective)
  ) +
  coord_fixed() +
  scale_fill_viridis_c()
```
