---
title: "Optimizing Multiple Responses"
description: |
  Tracking several criteria simultaneously.
author:
  - name: Kris Sankaran
    url: {}
date: 12-02-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = FALSE, fig.align = "center")
```

_Readings [11.3](https://www.wiley.com/en-us/Design+and+Analysis+of+Experiments%2C+10th+Edition-p-9781119492443#content-section), [Rmarkdown](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/2021-08-17-week12-2/week12-2.Rmd)_

```{r}
library(ggplot2)
theme424 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(theme424)
```

```{r}
opts_chunk$set(echo = TRUE)
```

1. In more complicated systems, we may want to optimize several objectives
simultaneously. More often than not, the goals will be at odds with one another.
For example, in our running chemical process example, we want to maximize yield
while maintaining a target viscosity and minimizing molecular weight. How can we
use response surface methods when we have several competing objectives?

### Overlaying Contours

2. The most direct approach is to simply fit several response surfaces. Then, we can visually inspect results to find factor configurations with desirable values across each response.

3. To illustrate, we will work with a chemical process dataset, which varied
temperature and time in some chemical process and observed changes in three
responses -- yield, viscosity, and molecular weight. The block below reads in
the data and defines a version, `mchem` that puts all three responses into one
column, using `pivot_longer`. We standardize all the responses in `mchem` so
that when we plot them below, the colors are all on the same scale.

```{r}
library(readr)
library(dplyr)
library(tidyr)

chem <- read_csv("https://uwmadison.box.com/shared/static/nbaj1m8j7tuaqmznjlrsgbzyhp9k61i8.csv")

mchem <- chem %>%
  mutate_at(vars(yield:molecular_weight), scale) %>% # standardize
  pivot_longer(yield:molecular_weight, names_to = "response")
```
4. The plot below shows how the three responses vary across the input
temperature and time. It seems that molecular weight is highest at high times
and temperatures. Viscosity is highest at average temperatures, and doesn't seem
to depend on time. Yield is highest at the center point and at high temperature
/ time.

```{r}
ggplot(mchem) +
  geom_point(
    aes(time, temp, col = value),
    position = position_jitter(w = 0.3, h = 0.3)
  ) +
  facet_grid(. ~ response) +
  scale_color_viridis_c()
```

5. We can improve this view by fitting a response model surface model to each
response. Then, we can view the estimated values across all times and
temperatures, not just the ones that were sampled. Below, we code the original
times and temperatures and fit a second order model associated with each run.

```{r}
library(rsm)

chem_coded <- coded.data(chem, time_coded ~ (time - 35) / 5, temp_coded ~ (temp - 155) / 5)
fits <- list(
  "yield" = rsm(yield ~ SO(temp_coded, time_coded), data = chem_coded),
  "viscosity" = rsm(viscosity ~ SO(temp_coded, time_coded), data = chem_coded),
  "molecular_weight" = rsm(molecular_weight ~ SO(temp_coded, time_coded), data = chem_coded)
)
```

6. Below, we visualize the contours associated with each of the three responses.
The conclusions are consistent with the earlier discussion. Moreover, they
highlight specific temperatures and times that would be optimal for each of the
responses. However, it's unclear how we would reconcile these three surfaces to
decide on a single temperature or time for the process.

```{r, fig.cap = "Three separate response surfaces, fit to yield, viscosity, and molecular weight, respectively."}
contour(fits[[1]], ~ time_coded + temp_coded, image = TRUE, asp = 1)
contour(fits[[2]], ~ time_coded + temp_coded, image = TRUE, asp = 1)
contour(fits[[3]], ~ time_coded + temp_coded, image = TRUE, asp = 1)
```


### Constrained Optimization

7. Whenever visual inspection is challenging, mathematical formalizations can
offer support. One idea is to frame the multiple response surface problem as a
constrained optimization.
  * Define acceptable ranges for responses  $y_{2}\left(x\right), \dots,
  y_{R}\left(x\right)$ that are important, but not our main focus.
  * Optimize the response $y_{1}\left(x\right)$ that’s our main focus.

8. Formally, we look for a configuration of factors $x_{\ast}$ that solves the
optimization
\begin{align*}
\underset{x}{\text{maximize}}\medspace &y_{1}\left(x\right) \\
\text{subject to }\medspace &\left(y_{2}\left(x\right), \dots, y_{R}\left(x\right)\right) \in \mathcal{C}
\end{align*}
where $C$ is the predefined acceptable region for the secondary responses.

### Desirability Functions

9. The main downside of the constrained optimization approach is that it forces
us to choose one response to prioritize over all others. What if we care about
each response more or less equally?

10. One idea is to optimize a sort of (geometric) averaged response,
\begin{align*}
\underset{x}{\text{maximize}}\medspace \left[\prod_{r = 1}^{R} y_{r}\left(x\right)\right]^{\frac{1}{R}}
\end{align*}

11. The issue with this idea is that it treats all responses exactly equally.
What if we want to maximize some, but minimize others? What if we want some to
be near some target value?

12. The solution is to use _desirability functions_. A few are plotted below. You
can adjust their shape so that the $r^{th}$ desirability function is large for
the values of $y_{r}\left(x\right)$ which are good (sloping down when you want
to minimize, sloping up when you want to maximize).

```{r, fig.cap = "Example desirability functions, for maximizing, minimizing, and achieving a target response.", echo = FALSE}
library(stringr)
library(desirability)

x_grid <- seq(-1, 1, .01)
example_funs <- data.frame(
  x = x_grid,
  d_max_0.5 = predict(dMax(0, 1, 0.5),  x_grid),
  d_max_1 = predict(dMax(0, 1, 1),  x_grid),
  d_max_2 = predict(dMax(0, 1, 2), x_grid),
  d_min_0.5 = predict(dMin(-1, 0, 0.5), x_grid),
  d_min_1 = predict(dMin(-1, 0, 1), x_grid),
  d_min_2 = predict(dMin(-1, 0, 2), x_grid),
  d_target_0.5 = predict(dTarget(-1, 0, 1, 0.5, 0.5), x_grid),
  d_target_1 = predict(dTarget(-1, 0, 1), x_grid),
  d_target_2 = predict(dTarget(-1, 0, 1, 2, 2), x_grid)
) %>%
  pivot_longer(-x, names_to="fun") %>%
  mutate(
    type = str_replace(fun, "[\\.0-9]+", ""),
    scale = str_extract(fun, "[\\.0-9]+")
  )
```


```{r, fig.cap = "Example desirability functions, for maximizing, minimizing, and achieving a target response.", echo = FALSE}
ggplot(example_funs) +
  geom_line(aes(x, value, col = scale)) +
  scale_color_brewer(palette = "Set2") +
  facet_grid(type ~ .)
```

13. Then, instead of optimizing the raw averaged response, we optimize the
averaged response after first passing through the desirability functions,
\begin{align*}
\underset{x}{\text{maximize}} \medspace \left[\prod_{r = 1}^{R} d_{r}\left(y_{r}\left(x\right)\right)\right]^{\frac{1}{R}}
\end{align*}

### Data Example

14. For the chemical problem, we can define a desirability function per
response, and then combine them into an overall objective. This is done using
the `desirability` package.

```{r, echo = TRUE}
library(desirability)
d_yield <- dMax(70, 95) # min / max / scale
d_viscosity <- dTarget(55, 65, 75) # min / target / max
d_weight <- dMin(2750, 4000) # min / max
objective <- dOverall(d_yield, d_viscosity, d_weight)
```

15. Now, we can apply this objective function to predictions made by the three
response surface fits we made above, in the section on overlaying contours.
We're evaluating the predictions over a grid of values in the range of the coded
time and temperature.

```{r}
coded <- as.data.frame(chem_coded)
x_grid <- expand.grid(
  time_coded = seq(min(coded$time_coded), max(coded$time_coded), .1),
  temp_coded = seq(min(coded$temp_coded), max(coded$temp_coded), 0.1)
)
```

```{r}
desirabilities <- cbind(
  x_grid,
  y1 = predict(fits[[1]], x_grid),
  y2 = predict(fits[[2]], x_grid),
  y3 = predict(fits[[3]], x_grid)
)
y_hat <- desirabilities %>% 
  select(starts_with("y"))
desirabilities$objective <- predict(objective, y_hat)
```

16. At this point, we can plot the desirability of each point in the time /
temperature configuration space. Note that a second mode emerges along the
low-temperature region -- these are regions which have good viscosity and
molecular weight properties, even though their yield isn't as high. Try
increasing the scale for yield to see what happens when you make the requirement
for high yield more stringent.

```{r, fig.margin = FALSE, fig.cap = "Overall desirability, considering yield, viscosity, and molecular weight."}
ggplot(desirabilities) +
  geom_tile(
    aes(time_coded, temp_coded, fill = objective)
  ) +
  coord_fixed() +
  scale_fill_viridis_c()
```
