---
title: "Unreplicated $2^K$ Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 11-02-2021
output:
  distill::distill_article:
    self_contained: false
---

_Readings [6.5](), [Rmarkdown]()_


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```


```{r}
library(BsMD)
library(dplyr)
library(ggplot2)
library(readr)
theme_set(theme_bw())
```

```{r}
# helper functions
code <- function(x) ifelse(x == '-', -1, 1)
daniel_plot <- function(effects) { 
  qq <- qqnorm(effects, datax = TRUE)
  qqline(effects, col = "red", probs = c(0.3, 0.7), datax = TRUE)
  text(qq$x, qq$y, names(effects), pos=4)
}
```


* Sometimes, people will only take only one measurement per factor configuration

	* When $K$ is large, replication can increase the number of samples needed
	substantially
	
	*  E.g., changing n from 1 to 2 when K = 5 means 32 more runs
	
* For factor screening experiments, typically want to allocate samples to
understanding new factors, rather than replicating known configurations

Without replicates to gauge measurement noise, we may encounter two opposite problems,

* Missing a true effect
* Spurious effects

Let’s see how these problems arise, and discuss some solutions.

## Missing True Effects

* If the effect is weak, then if only nearby levels are tested, the effect will
be easy to miss

* Fix: We can space out the levels at which we test each factor
```{r, fig.margin = TRUE}
library("EBImage")
display(readImage("https://uwmadison.box.com/shared/static/olbjp5fubt4mm6j1fa7rthe0uoctymef.pnghttps://uwmadison.box.com/shared/static/8h0vt5kcb5ss0l9dlghsswnpf88q8267.png"))
display(readImage("https://uwmadison.box.com/shared/static/099eopgmbncci9ob9zjo1mvvguxinibn.png"))
```

## Spurious Effects

* If there are no replicates, then we can perfectly interpolate the data

* Leaves us with no degrees-of-freedom for estimating $\sigma^2$

	* $\sigma^2$ is needed to perform ANOVA
	
```{r, fig.margin = TRUE}
display(readImage("https://uwmadison.box.com/shared/static/fcujd4j9ano0ypij18w9jy70amtr00p3.png"))
display(readImage("https://uwmadison.box.com/shared/static/rfdz734pv6rxolgwz5fx215qb41qm0k6.png"))
```
	
* Fix: Pool together higher-order interactions

	* High-order interactions are typically rare (sparsity of effects principle)
	
	* Can use pooled interaction estimates to obtain $\sigma^2$
	
	* Pooling can only make testing more conservative
	
```{r, fig.margin = TRUE, fig.height = 4}
filtration <- read_table2("https://uwmadison.box.com/shared/static/xxh05ngikmscnddbhg2l3v268jnu4jtc.txt") %>%
  mutate_at(vars(-Rate), as.factor)
ggplot(filtration) +
  geom_point(aes(x = A, y = Rate, col = C)) +
  scale_color_brewer(palette = "Set2") +
  facet_grid(B ~ D)
```

## Data Example

This is what happens when we try to estimate all interactions, even though there
is only one replicate per factor configuration. We'll use the filtration
dataset, which asks how temperature (A), pressure (B), formaldehyde (C), and
stirring rate (D) affect the filtration rate of the resulting product. 

```{r}
fit <- lm(Rate ~ A * B * C * D, data = filtration)
summary(fit)
anova(fit)
```
If we instead assume that all interactions terms of order 3 or higher are null,
we can perform again perform ANOVA.

```{r}
fit <- lm(Rate ~ (A + B + C + D) ^ 2, data = filtration)
summary(fit)
anova(fit)
```

## Design Projections

* Related to pooling, sometimes it is clear that a certain factor has no bearing
on the response

* In this case, we may consider removing that factor and all the interaction
terms that include it

* By collapsing the factor, we double the number of effective replicates
```{r, fig.margin = TRUE}
display(readImage("https://uwmadison.box.com/shared/static/yhc2w1698namew8r0vgn7ej7546ev2wg.png"))
```

## Visualing Effects

Graphical methods provide useful summaries for evaluating interactions in the $2^K$ model.

* Daniel Plots

  * If none of the factors had any influence on the response, then the effects
  would all be normally distributed around 0.
  
  * Idea: make a normal probability plot of the effects, and look for those which
  deviate from identity line. These are likely real effects.
  
```{r, fig.margin = TRUE, fig.height = 5}
# estimate effects
filtration_coded <- filtration %>%
  mutate_at(vars(-Rate), code)
fit_coded <- lm(Rate~A*B*C*D, data=filtration_coded)
effects <- 2 * coef(fit_coded)[-1] # exclude intercept
daniel_plot(effects)
```

* Lenth Plots

  * An alternative is to simply plot the effect sizes directly
  
```{r, fig.margin = TRUE, fig.height = 5}
LenthPlot(fit_coded, cex.fac = 0.4)
```

Let $s_0 = 1.5 \times \text{median}\left(\text{Contrast}_j\right)$.  Then, the
notation here refers to

* Pseudostandard error (PSE): $1.5 \times \text{median}\left(\left|c_j\right| :
\left|c_j\right| < 2.5 s_0\right)$ serves as an alternative to the usual
standard error over contrasts, which is robust to outliers (it completely
ignores $c_j$'s that are larger than $2.5 s_0$.)
* Margin of error (ME): A version of the critical $t$-value that relies on the
robust standard error. Defined as $t_{0.025,\frac{m}{3}}\times PSE$, where $m$
is the total number of effect estimates (columns in the Lenth plot).
* Simultaneous margin of error (SME): A more conservative version of ME, to
protect against multiple comparisons. 