---
title: "Nested Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-17-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [14.1](), [Rmarkdown]()_


```{r}
library(dplyr)
library(ggplot2)
library(lme4)
library(readr)
library(EBImage)
library(reshape2)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r, fig.cap = "The nested structure in the suppliers example"}
#display(readImage("https://uwmadison.box.com/shared/static/kbs80435ds41u0qp7typvmuiaibyjlpm.png"))
```

Factorial designs cover the situation where effects from multiple sources are
crossed with one another. In many situations, however, effects are nested within
each other.

* You rely on several suppliers to provide raw material and are interested in
whether some suppliers provide consistently better material. The material
arrives in batches, and there is variation across batches. The batch effect is
*nested within* the supplier effect.

* The hardness of different alloys are to be compared. Several ingots are built
from each alloy, and repeated measures are drawn from each ingot. The ingot
effect is nested within the alloy effect.

### Model

```{r, fig.cap = "Notation used in nested designs. Parentheses denote the parent branch."}
#display(readImage("https://uwmadison.box.com/shared/static/g6yb4uu8nncnqc39rbjhh3kz8bv9jzwc.png"))
```

We will consider the two-stage design here. The data are imagined to be drawn
according to
\begin{align*}
y_{ijk} &= \mu + \alpha_{i} + \beta_{j\left(i\right)} + \epsilon_{ijk}.
\end{align*}

```{r, fig.cap = "Another way to look at the nested coefficient structure.", out.width = "80%"}
#display(readImage("https://uwmadison.box.com/shared/static/fy3jwyeou1bu6nyo1i2lypg8juj80z8k.png"))
```

* $\alpha_i$ is the parent-effect (e.g., supplier effect). For identifiability,
assume $\sum_{i} \alpha_i = 0$.

* $\beta_{j\left(i\right)}$ is the nested-effect associated with parent $i$
(e.g., effect of the $j^{th}$ batch within the $i^{th}$ supplier). For
identifiability, assume $\sum_{j}\beta_{j\left(i\right)} = 0$ for each $i$.

* $\epsilon_{ijk} \sim N\left(0, \sigma^{2}\right)$ is independent noise.

The key difference from the usual factorial model is that the nested effects
$\beta_{j\left(i\right)}$ are not the same across different parents $i$.

### Inference

_Fixed-Effects_

First, imagine treating all the effects as fixed. In this case, there are two
typical hypotheses of interest.

```{r, fig.cap = "The nested effect term looks at the variation across child means along a single branch."}
#display(readImage("https://uwmadison.box.com/shared/static/6bona7dg8dl6fbnfvetovq2cl9l5zo11.png"))
```

* Null parent-effect,
\begin{align*}
&H_{0}: \alpha_{i} = 0 \text{ for all } i \\
&H_{1}: \alpha_{i} \neq 0 \text{ for at least one } i
\end{align*}
	
* Null child-effect,
\begin{align*}
&H_{0}: \beta_{j\left(i\right)} = 0 \text{ for all } i,j \\
&H_{1}: \beta_{j\left(i\right)} \neq 0 \text{ for at least one } i,j
\end{align*}

In either case, a test is performed using a sum-of-squares decomposition,
\begin{align*}
SS_{T} &= SS_A + SS_{B\left(A\right)} + SS_{E}
\end{align*}
which is similar in structure to those we have seen before, except we now have a nested effect term,
\begin{align*}
SS_{B\left(A\right)} &= \sum_{i = 1}^{a}\sum_{j = 1}^{b} \left(\bar{y}_{ij\cdot} - \bar{y}_{i\cdot\cdot}\right)^{2}.
\end{align*}

which measures how much the $j^{th}$ effect within parent $i$ varies from
the average in that group. The distribution of the associated mean-squares can
then be used to perform an ANOVA. Let’s see an implementation, based on the
supplier-materials example above. (example 14.1)

```{r}
purity <- read_csv("https://uwmadison.box.com/shared/static/uub0t6lvii52rxyygb2yt4ph3vmionjz.csv") %>%
  mutate(
    supplier = as.factor(supplier),
    batch = as.factor(batch)
  )
```
```{r, fig.cap = "Data from the supplier example."}
ggplot(purity) +
  geom_point( aes(x = batch, y = purity) ) +
  facet_wrap(~supplier)
```
```{r, echo = TRUE}
fit <- lm(purity ~ supplier + supplier / batch, data = purity)
anova(fit)
```
The `A/B` notation is just shorthand for `A + A:B`; i.e., a main effect for the
parent `A` and an interaction term between the parent `A` and child `B`, as the
code below makes clear.

```{r}
fit_explicit <- lm(purity ~ supplier + supplier : batch, data = purity)
anova(fit_explicit)
```
_Random Effects_

Rather than caring about specific parent or nested effects, we may simply want
to know the typical variation due to either factor. For example, we may not care
about the effect of the 2nd batch in the 3rd supplier, but we may be curious
about the typical size of batch-to-batch variation. In this case, it makes sense
to use random effects. We can use a random effect for just the nested factors,

```{r, fig.margin = FALSE, fig.cap = "We might imagine the batch effects are drawn from some distribution, and do inference on the variance of that distribution."}
#display(readImage("https://uwmadison.box.com/shared/static/twey7ufa1zeaq8lgjpdg1uu66a7fi2l4.png"))
```


```{r}
fit <- aov(purity ~ supplier + Error(supplier/batch), data = purity)
summary(fit)
```
or for both the nested and parent factors,
```{r}
fit <- lmer(purity ~ (1 | supplier / batch), data = purity)
summary(fit)
```
However, it makes no sense to treat the parent as random, but its child as
fixed.

An important detail to keep in mind is that, for nested designs, computation of
$F$-statistics depends on which terms are treated as fixed or random. For
example, suppose $B$ is nested within $A$. Then the $F$ statistic for $A$ is
computed as

* $\frac{MS_A}{MS_E}$ if both $A$ and $B$ are fixed

* $\frac{MS_A}{MS_B(A)}$ if $A$ is fixed but $B$ is random

* $\frac{MS_{B\left(A\right)}}{MS_{E}}$ if both $A$ and $B$ are random