---
title: "Experimental Design R Cheatsheet"
output:
  pdf_document:
    toc: true
    toc_depth: 2
urlcolor: magenta
---
```{r, echo = FALSE}
library(knitr)
opts_chunk$set(size="footnotesize", cache = TRUE, fig.width = 6, fig.height = 3, dpi = 200, echo = TRUE)
options(width = 300)
```

# Format

This cheatsheet includes R recipes that are useful for experimental design. The general format is,

## `function_name`

_package (if necessary)_

Brief description

```{r}
# code_example
```
I welcome you to leave comments or propose changes to the document on [github](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/r_cheatsheet.Rmd),
using either an issue or pull request.

# Chapter 2

## `mean`

Computes the sample mean $\frac{1}{n}\sum_{i = 1}^{n} x_i$ of a vector of $n$ numeric
values.
```{r}
mean(1:10)
```

## `sd`

Computes the sample standard deviation $\frac{1}{n - 1}\sum_{i = 1}^{n}\left(x_i - \bar{x}\right)^2$ of a vector of $n$ numeric values.
```{r}
sd(1:10)
```

## `rnorm`, `rt`, `rchisq`, `rf`

Simulates $n$ samples from a normal, $t$, chi-square, or $F$-distribution,
respectively. The parameters of the normal are the mean and standard deviation.
The other distributions require degrees-of-freedom parameters.
```{r}
n <- 5
rnorm(n, 0, 1)
rt(n, 1)
rchisq(n, 1)
rf(n, 1, 2)
```

## `pnorm`, `pt`, `pchisq`, `pf`

Computes the total probability of landing below $x$ in a normal, $t$,
chi-square, or $F$-distribution, respectively. Accepts the same parameters as the `r`-prefixed functions above.
```{r}
x <- -1.96
pnorm(x, 0, 1)
pt(x, 1)

x <- 10
pchisq(x, 1)
pt(x, 1)
pf(x, 1, 2)
```


## `qnorm`, `qt`, `qchisq`, `qf`

Computes the $x$-coordinate such that a probability $q$ of the distribution lies
below $x$. Especially useful for computing cutoff values in hypothesis tests.

```{r}
q <- 0.05 / 2
qnorm(q)
qt(q, 1)
qchisq(q, 1)
qf(q, 1, 2)
```


## `dnorm`, `dt`, `dchisq`, `df`

Computes the probability density of a distribution at a specific value $x$.

```{r}
x <- 2
dnorm(x)
dt(x, 1)
dchisq(x, 1)
df(x, 1, 2)
```


## `z.test`

_`library(BSDA)`_

Performs a one or two-sample `z`-test when the variance of the data is known.
Returns the estimate of the (group) means, the test statistic, a $p$-value, and
a confidence interval (whose level is specified by the `conf.level` parameter).

```{r}
library(BSDA)
one_sample <- z.test(rnorm(5, 1), mu = 0, sigma.x = 1)
two_sample <- z.test(rnorm(5), rnorm(5, 1), sigma.x = 1, sigma.y = 1)
two_sample
two_sample$p.value # p.value
two_sample$conf.int # confidence interval for the difference
```

## `t.test`

Performs a one or two-sample `t`-test, testing when the variance is unknown. We
typically assume equal variance (`var.equal = TRUE`) when testing for a
difference in means unless an initial / diagnostic analysis explicitly suggests
to do otherwise. Returns the estimate of the means, the test statistic, a
$p$-value, and a confidence interval (whose level is specified by the
`conf.level` parameter).

```{r}
one_sample <- t.test(rnorm(10, 1), mu = 0)
two_sample <- t.test(rnorm(10), rnorm(10, 1), var.equal = TRUE)
two_sample
two_sample$p.value
two_sample$conf.int
```

## `hist`

Built-in function to make histograms in R.

```{r}
hist(rnorm(10))
```


## `geom_histogram`

_`library(ggplot2)`_

`ggplot2` alternative to making histograms. The data must be input as a
data.frame.

```{r}
library(ggplot2)
df <- data.frame(x = rnorm(10))
ggplot(df) +
  geom_histogram(aes(x))
```

## `plot(x, type = "l")`

Base R approach to generating a line plot.

```{r}
plot(rnorm(10), type = "l")
```

## `geom_line`

_`library(ggplot2)`_

`ggplot2` approach to generating a line plot. Expects a `data.frame` with at
least two columns as input, used to specify the $x$ and $y$ coordinates.

```{r}
library(ggplot2)
df <- data.frame(x = 1:10, y = rnorm(10))
ggplot(df) +
  geom_line(aes(x, y))
```

# Chapter 3

## `lm` and `aov`

Compute an ANOVA table describing how multiple levels of a factor are related to
a response. Expects as input a `data.frame` whose columns include the response
values and factor levels. Beware that if the levels are written as numeric
values, then R will not treat each value as a separate group (use `as.factor` to
convert into discrete groups). Note that the `anova()` function returns
comparable output, but is not well-suited to contrast calculations.

```{r}
experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

fit <- lm(y ~ levels, data = experiment)
aov_table <- aov(fit)
summary(aov_table) # print the test results
```
## `tidy`

_`library(broom)`_

Convert the results from a base R test into a data.frame. Useful for extracting
test results.

```{r}
library(broom)
aov_df <- tidy(aov_table) # uses result from previous example
aov_df
aov_df$p.value
```

## `resid`

Extracts the residuals $e_{ij} = y_{ij} - \hat{y}_{ij}$ associated with a model. This is useful for checking model
assumptions.

```{r}
experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

fit <- lm(y ~ levels, data = experiment)
resid(fit)
```

## `predict`

Extracts the fitted values $\hat{y}_{ij} = \hat{\mu} + \hat{\tau}_i$ associated
with a model. Often useful when performing model checks.

```{r}
experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

fit <- lm(y ~ levels, data = experiment)
predict(fit)
```

## `qqnorm` and `qqline`

Used to make QQ plots against a theoretical normal distribution. Usuall applied
to the residuals of a fitted model.

```{r}
qqnorm(resid(fit))
qqline(resid(fit), col = "red")
```

## `pivot_longer`

_`library(tidyr)`_

Used to transform a "wide" dataset into a "tall" one. This is often useful for
moving the replicates for an experimental factor into a single column, so that
it can be used by the `lm` function.

```{r}
library(tidyr)

experiment <- data.frame(
  levels = c("A", "B", "C"),
  rep1 = rnorm(3),
  rep2 = rnorm(3)
)
experiment_tall <- pivot_longer(experiment, -levels, names_to = "replicate")

experiment
experiment_tall
```

## `geom_point`

_`library(ggplot2)`_

`ggplot2` approach to drawing scatterplots. Often useful for plotting response
values against a factor of interest.

```{r}
experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

ggplot(experiment) +
  geom_point(aes(levels, y))
```

## `fit.contrast`

_`library(gmodels)`_

Conduct a hypothesis test for a specific contrast ($H_0: \Gamma\left(c\right) =
0$ in an ANOVA model. Can be applied to multiple contrasts simultaneously by
passing the vectors $c_1, \dots, c_m$ as rows of a matrix.


```{r}
library(gmodels)

experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

# fit an anova model -- see entry on `aov` and `lm`
fit <- lm(y ~ levels, data = experiment)
aov_table <- aov(fit)

# contrast between levels A and B
contrast <- c(1, -1, 0)
fit.contrast(aov_table, "levels", contrast)

# several contrasts
contrasts <- matrix(c(1, -1, 0, 1, 1, -1),  nrow = 2, byrow = TRUE)
fit.contrast(aov_table, "levels", contrasts)
```

## `fit.contrast(..., conf.int = T)`

_`library(gmodels)`_

Compute the confidence interval for a contrast. Used in the exact same way as
`fit.contrast`, but setting the `conf.int` parameter to the desired confidence
level.

```{r}
# continues from previous example
fit.contrast(aov_table, "levels", contrast, conf.int = 0.95)
```
## `pairwise.t.test`

Tests for equality in means between all pairs of levels of a factor, without any
multiple comparison adjustment. The first argument is a vector of response
values, the second is a vector containing the group membership. Results are
typically similar to Fisher's LSD, unless some of the groups are very small.

```{r}
experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

pairwise.t.test(experiment$y, experiment$levels)
```


## `PostHocTest(..., method = "scheffe")`

_`library(DescTools)`_

Adjust the widths of confidence intervals for multiple contrasts using Scheffe's
method.

```{r}
library(DescTools)

experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

# fit an anova model -- see entry on `aov` and `lm`
fit <- lm(y ~ levels, data = experiment)
aov_table <- aov(fit)

# two contrasts
contrasts <- matrix(c(1, -1, 0, 1, 1, -1),  nrow = 2, byrow = TRUE)
PostHocTest(aov_table, method = "scheffe", contrast = t(contrasts))
```

## `TukeyHSD`

_`library(DescTools)`_

Use Tukey's Honest Significant Difference method to build experimentwise-valid
confidence intervals for the contrast between all pairs of factor levels.

```{r}
library(DescTools)
experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

# fit an anova model -- see entry on `aov` and `lm`
aov_table <- aov(lm(y ~ levels, data = experiment))
TukeyHSD(aov_table)
```

## `PostHocTest(..., method = "lsd")`

_`library(DescTools)`_

Use Fisher's Least Significance Difference method to perform tests for the
differences between all pairs of factor levels. Note that Fisher's method does
not control the experimentwise error rate, like Tukey's or Scheffe's methods do.

```{r}
library(DescTools)

experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

# fit an anova model -- see entry on `aov` and `lm`
fit <- lm(y ~ levels, data = experiment)
aov_table <- aov(fit)

# two contrasts
PostHocTest(aov_table, method = "lsd")
```

## `lmer`

_`library(lme4)`_

Fits a random effects model using the restricted maximum likelihood method. The
syntax is the same as for `lm`, except that random effects terms can be
specified using `(1 | random_term)`.

```{r}
library(lme4)

experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

fit <- lmer(y ~ (1 | levels), data = experiment)
fit
```
## `confint`

Compute confidence intervals for $\sigma_{\tau}$ and $\sigma$ in a random
effects model fitted using the restricted maximum likelihood method. Note that
to get confidence intervals on $\sigma_{\tau}^2$ and $\sigma^2$, both endpoints
of these intervals have to be squared.

```{r}
confint(fit) # continuation of previous example
```

## `kruskal.test`

Performs a nonparametric ANOVA test, based on ranks rather than any assumed
probability model. The input syntax is exactly the same as `lm` for ordinary
ANOVA.

```{r}
experiment <- data.frame(
  levels = rep(c("A", "B", "C"), each = 5),
  y = rnorm(15)
)

fit <- kruskal.test(y ~ levels, data = experiment)
fit
```

# Chapter 4

## `BIBsize`

_`library(daewr)`_

Uses the identity $\lambda (a - 1) = r(k - 1)$ from BIBD designs to determine
possible balanced designs when fixing two of the values among $a, r, k$. Can be
helpful when trying to see whether a BIBD can be applied under certain
experimental contraints (like, the maximum number of samples that can be taken
from each block).

```{r}
library(daewr)
BIBsize(t = 5, k = 3) # we should use 10 blocks
```

## `find.BIB`

_`library(crossdes)`_

Given the three parameters $a, r, k$ from a BIBD, comes up with an assignment of
treatments to blocks in a way that treatments are balanced across all blocks.
Often used in conjunction with `BIBsize`.

```{r}
library(crossdes)
find.BIB(trt = 5, k = 3, b = 10)
```

