---
title: "Saturated Designs"
description: |
  Very efficient analysis of a large set of factors.
author:
  - name: Kris Sankaran
    url: {}
date: 11-23-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = FALSE)
```

_Readings [8.6, 8.7](https://www.wiley.com/en-us/Design+and+Analysis+of+Experiments%2C+10th+Edition-p-9781119492443#content-section), [Rmarkdown](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/2021-08-17-week11-4/week11-4.Rmd)_

```{r}
library(ggplot2)
theme424 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(theme424)
```

1. There are designs that let you study $K$ factors using only $N + 1$ samples;
such designs are called _saturated_. This is an extremely efficient use of
samples.

2. The reason we don’t use saturated designs all the time are that (1) they are
not available for all choices of $K$ and (2) the resulting aliasing structure
can make definitive inferences difficult. That said, it’s worth being familiar
with a few saturated designs, since it can result in dramatically reduced
sampling effort in some special cases.

### Resolution III Designs

3. When $K + 1$ is a power of 2, then there are resolution III designs that use
only $N = K + 1$ samples. Here are some examples,

* $2^{3 - 1}_{III}$: 3 factors in 4 samples
  - Generators: $C = AB$

| A | B | C |
| --- | --- | --- |
| - | - | + |
| - |  + | - |
| +| - | - |
| + | + | + |

* $2^{7 - 4}_{III}$: 7 factors in 8 samples

	* Generators: $D = AB, E = AC, F = BC, G = ABC$
	* See Table VIII(h) for alias groups.
	
* $2^{15 - 11}_{III}$: 15 factors in 16 samples

	* See Table VII(h) for generators and alias groups (there are many)
	
* Exercise: Give another example (e.g., for 31 factors?)

### Plackett-Burman Designs

4. Plackett-Burman designs are a collection of design options working outside
the usual fractional factorial paradigm — they don’t rely on the ideas of
generators or defining relations that we’ve been using so far to subset full
factorial designs into less-costly experiments.

5. We won’t describe their construction, which relies on techniques from
abstract algebra. The properties that are most important to know are
  * A saturated design can be found whenever $K + 1$ is a multiple of 4, which
  is much more often than in the resolution III situation above
  * Their aliasing structure is typically more complicated. For example, when $K
  = 11$, then every main effect is aliased with every pairwise interaction not
  including that main effect
  * Their projectivity properties are typically good.
  	* E.g., resolution III Plackett-Burman has projectivity 3, while any $2^{k -
  	p}_{III}$  fractional factorial has only projectivity 2
  	* So, we can collapse to any subset of 3 factors and get a full factorial
	
6. To use these designs, you have to look up the tables of +/-’s, saying which
factors to keep on or off for each run. See e.g., Table 8.23/8.24.
Alternatively, use the `FrF2` R package.