---
title: "Blocking in Response Surface Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-17-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [11.4](), [Rmarkdown]()_



```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(EBImage)
library(reshape2)
library(rsm)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r, fig.cap = "By tying sampling center points in both the axial and factorial blocks, it becomes possible to estimate and correct for block effects."}
#display(readImage("https://uwmadison.box.com/shared/static/hxsz37wvxcj93osymlckvep6da6ipoqy.png"))
```

It’s often impossible to collect all the samples for an experiment at once,
which can lead to batch effects downstream. This motivates the use of blocked
designs — how should we think of blocked designs in the context of response
surface experiments?

First, let’s build intuition through an example. Suppose we are using a central
composite design and $K = 2$, but we can’t complete all 10 runs at once. At
most, during a single pass, we can collect 5 samples.

Idea: Split the factorial and axial samples into two blocks, but tie them
together at the center points.

* Any batch effects can be directly estimated through shifts in the average
value at the center

The main technical condition that generalizes this idea is called _block
orthogonality_. A design is called block orthogonal if,

* Each block $X_{b}$ is orthogonal, i.e., $X_{b}^{T}X_{b}$ = I when we stack
all design points for the block into an $n_{b} \times K$ matrix $X_{b}$.

* For each factor $k$ and each block $b$,
\begin{align*}
  \frac{\sum_{u = 1}^{n_{b}} x_{iu}^{2}}{\sum_{u = 1}^{N} x_{iu}^{2}} = \frac{n_{b}}{N}
\end{align*}
  this ensures that no one block has undue influence over the eventual estimates.
	
For arbitrary number of factors $K$, the central composite design turns can be
made block orthogonal by dividing into two blocks (factor and axial points, tied
at the center). It can be broken into even more blocks for particular choices of
block size; this is summarized in Table 11.11.

### Code Example

We can use the same code we used to generate central composite designs before,
but using the `blocks` argument.

```{r}
codings <- list(
  time_coded ~ (time - 35) / 5,
  temp_coded ~ (temp - 150) / 5
)
ccd_design <- ccd(~ time_coded + temp_coded, coding = codings)
ccd_design
```

```{r}
ggplot(decode.data(ccd_design)) +
  geom_point(
    aes(x = time, y = temp, col = Block),
    position = position_jitter(w = 0.5, h = 0.5),
    size = 3
  ) +
  scale_color_brewer(palette = "Set2") + 
  coord_fixed()
```

Suppose we wanted to block more than just the factorial and the axial runs; for
example, we may only be able to run 2 examples per batch.  In this case, we
might block the factorial runs by confounding a high-order interaction with the
blocks. For example, the code below blocks the 32 runs of a $2^{5}$ factorial
into blocks of size 8.

```{r, echo = TRUE}
codings <- list(
  time_ ~ (time - 35) / 5,
  temp_ ~ (temp - 150) / 5,
  power_ ~ power,
  rate_ ~ rate,
  cooling_ ~ cooling
)
blocked_ccd <- ccd(~ time_ + temp_ + power_ + rate_ + cooling_, coding = codings, blocks = Block ~ c(time_ * temp_ * power_, power_ * rate_ * cooling_))
```

Finally, let's check the orthogonal blocking property for one of the blocks.
Remember, we need to make sure that (1) the design within the block is diagonal
and (2) the fraction of norm for a column is proportional to the number of
samples within the block. This is checked below for Block 1.

```{r}
ggplot(decode.data(blocked_ccd)) +
  geom_point(
    aes(x = time, y = temp, col = Block),
    position = position_jitter(w = 1, h = 1),
    size = 3, alpha = 0.6
  ) +
  scale_color_brewer(palette = "Set2") + 
  coord_fixed()
```

```{r, echo = TRUE, size = "scriptsize"}
x <- as.data.frame(blocked_ccd)
xb <- x[x$Block == 1, ] %>%
  select(ends_with("_")) %>%
  as.matrix()
t(xb) %*% xb # orthogonality
sum(xb[, 1]^2) / sum(x$time ^ 2) # ||xb[1]||^2 / ||x[1]||^2
12 / nrow(blocked_ccd) # nb / N
```
