---
title: "$2^{K - 1}$ Fractional Factorial Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-17-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [8.1 - 8.2](), [Rmarkdown]()_


```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(EBImage)
library(ggplot2)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

```{r}
# helper functions
code <- function(x) ifelse(x == '-', -1, 1)
daniel_plot <- function(effects, probs = c(0.4, 0.6)) { 
  qq <- qqnorm(effects, datax = TRUE)
  qqline(effects, col = "red", probs = probs, datax = TRUE)
  text(qq$x, qq$y, names(effects), pos=1)
}
```


The trouble with $2^{K}$ designs is that the number of samples required grows
exponentially with the number of factors $K$. The differences between 4 and 7
factors doesn’t sound like a lot, but it’s the difference between 16 and 128
experimental runs. Fractional factorial designs show how to study extra factors
without having to make as many runs. There is no free lunch, but the tradeoffs
that it creates are often worth it.

### Principles

There are two central ideas,

* Sparsity of Effects: It’s often safe to assume that high-order interactions
are rare.

* Heredity: It is unusual for an interaction to be significant without the
corresponding main effects also being significant. For example, it would be
unusual to have $A \times B$ be significant without at least one of $A$ and $B$
being significant.

### Initial Exploration

```{r, fig.cap = "A bad choice of samples to subset down to, which would make it impossible to estimate effects from $C$."}
display(readImage("https://uwmadison.box.com/shared/static/gu87vet64i970eonsbvtds7bsb31qbuy.png"))
```
Suppose we want to study $A, B$ and $C$, but we only have a budget of 4 runs. We
need to choose four corners of the cube to sample. Some choices are obviously
bad. E.g., if we chose all the samples on one face, we wouldn’t see any
variation in one factor. Here’s an idea that will lead us on the right path.
Recall our old notation,

| label | A | B |  AB |
|--- | --- | --- | --- |
| (1) | - | - | + |
| a | + | - | - |
| b | - | + | - |
| ab | + | + | + |

Acting on the sparsity of effects principle, let’s suppose that AB is null and
invent a contrast for C that exactly matches AB. We’ll update the label to match
the associated corner of the cube.

| label | A | B |  AB | C |
|---| --- | --- | --- | --- |
| (c) | - | - | + | + |
| a | + | - | - | - |
| b | - | + | - | - |
| abc | + | + | + | + |

```{r, fig.cap="A more sensible fraction of samples to take."}
display(readImage("https://uwmadison.box.com/shared/static/mmj9mcixl4y0o95hwf7h93vayw0eooc0.png"))
```

This seems like a reasonable choice of 4 corners, making it possible to estimate
at least all the main effects. This turns out to be a general strategy for
subsetting the full design. But before we do another example, we need to define
some vocabulary.

### Vocabulary

* $2^{K - 1}$: This denotes a fractional factorial design where we take $\frac{1}{2}$ of the samples in a $2^{K}$ design.

* Aliases: Above, we assumed that AB was null. Suppose that it weren’t, though.
Notice that the contrasts used for the pairs below are all equal,
\begin{align*}
	 \left[AB\right] &= \frac{1}{2}\left(c - a + b + abc\right) = \left[C\right] \\
		\left[BC\right] &= \frac{1}{2}\left(-c + a - b + abc\right) = \left[A\right] \\
		\left[AC\right] &= \frac{1}{2}\left(-c - a + b + abc\right) = \left[B\right]
\end{align*}

and we have no conclusive way of distinguishing between these *aliased* effects,
besides appeals to the hereditary principle or domain knowledge. To denote this
unidentifiability, we will use bracket notation, for example,
\begin{align*}
\left[A\right] &= A + BC.
\end{align*}

* Generators: In the example, we set $C = AB$. Multiplying both sides by $C$ and
using the fact that $C^2 = I$, we find $ABC = I$. The *word* ABC will be called
the generator for this fraction. 

```{r, fig.cap = "An example complementary fraction."}
#display(readImage("https://uwmadison.box.com/shared/static/pffnjyav7wexbujzf32vlhorcfxhfa6w.png"))
```

* Complementary designs: The complementary fraction of a fractional factorial
design are the corners on which we didn’t take samples. Often, if there are
strong aliased effects in an initial fractional design, the complementary
fraction will be run in the next experiment.

* Resolution: A fractional design has resolution $R$ if no $p$-factor is aliased
with an effect containing less than $R - p$ factors.

| $R$ | $p$ | $< R - p$ | Interpretation |
|- | - | --- | ------- |
| 3 | 1 |  $\leq 2$ | Main effects aren't aliased with other main effects, but could be aliased with two-way interactions | 
| 4 | 1 | $\leq 3$ | Main effects aren't aliased with any other main effects or with any two-way interactions, but could be aliased with three-way interactions | 
|  | 2 | $\leq 2$ | Two-way interactions aren’t aliased with main effects.
| 5 | 1 | $\leq 4$ | Main effects aren’t aliased with other main effects, two-way, or three way interactions. | 
| | 2 | $\leq 3$ | Two-way interactions aren’t aliased with with main effects or two-way interactions. |

* Projection: In a fractional factorial, if we decided that we don’t care about
a factor after all, we automatically end up with a $2 ^ {K - 1}$  full-factorial
for free.
```{r fig.cap="Three projections of $2^3$ designs into $2^2$ designs, from the textbook."}
#display(readImage("https://uwmadison.box.com/shared/static/b94hfeyugeun1xywlm2ccls9am5l38xg.jpg"))
```

* Sequences: If we find a strong contrast for an aliased effect, it’s natural to
try a follow-up experiment to resolve the ambiguity.

### Code Example

Let's suppose we only had only run half of the runs in the filtration example
from week 9 [1]. The original experiment is a $2^ 4$ design; we will choose the
fraction corresponding to the word^[The word $I = ABCD$ emerged from defining $D
= ABC$.] $I = ABCD$

This is a general principle for $2 ^ {K - 1}$ designs: choose which
configurations to run by defining the relation $K = A B C \dots J$. This turns a
full $2^{K - 1}$ factorial design, which only studied $K - 1$ factors into a
fractional $2^{K - 1}$ factorial that studies $K$ factors.

```{r, echo = FALSE}
filtration <- read_table2("https://uwmadison.box.com/shared/static/xxh05ngikmscnddbhg2l3v268jnu4jtc.txt") %>%
  mutate_at(vars(-Rate), as.factor) %>%
  mutate_at(vars(-Rate), code) %>%
  filter(D == A * B * C)
```
We can perform the fit. Note that there missing values in the estimated
coefficients because some effects are aliased.
```{r, echo = TRUE}
dim(filtration)
fit <- lm(Rate ~ A * B * C * D, data = filtration)
summary(fit)
```

To back out which terms are aliased, let's look carefully at the design matrix.
```{r, echo = TRUE, size = "scriptsize"}
X <- model.matrix(fit)
t(X) %*% X != 0 # TRUE on off diagonals are aliases
```
So, revisiting the estimated fit, we can conclude that.
\begin{align*}
\widehat{\mu}+\widehat{A B C D} &=70.75 \\
\widehat{A}+\widehat{B C D} &=9.5 \times 2=19 \\
\widehat{B}+\widehat{A C D} &=0.75 \times 2=1.5 \\
\widehat{C}+\widehat{A B D} &=7 \times 2=14 \\
\widehat{D}+\widehat{A B C} &=8.25 \times 2=16.5 \\
\widehat{A B}+\widehat{C D} &=-0.5 \times 2=-1 \\
\widehat{A C}+\widehat{B D}=&-9.25 \times 2=-18.5 \\
\widehat{B C}+\widehat{A D} &=9.5 \times 2=19
\end{align*}

Let's see which of the effects are important, and see whether we can draw any
reasonable conclusions, based on the hereditary principle.
```{r, echo = FALSE}
daniel_plot(2 * coef(fit)[-1], prob=c(0.25, 0.3))
```
The important effects look like,

* $A + BCD$ 

* $BC + AD$

* $D + ABC$

* $C + ABD$

* $AC + BD$

It seems like we should believe in the main effects $A, D$, and $C$. For the
two-way interactions, $AC$ and $AD$ are more plausible, because the main effect
for $B$ was not significant.