---
title: "General Factorial Designs"
description: |
  Factorial designs with arbitrary numbers of factors
author:
  - name: Kris Sankaran
    url: {}
date: 10-21-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [5.4](https://www.wiley.com/en-us/Design+and+Analysis+of+Experiments%2C+10th+Edition-p-9781119492443#content-section), [Rmarkdown](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/2021-08-16-week7-3/week7-3.Rmd)_

```{r, echo = FALSE}
library(dplyr)
library(ggplot2)
library(EBImage)
library(readr)
theme424 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(theme424)
```

1. We’ll discuss three factor factorial designs, with the hope that what we
learn will generalize to arbitrary numbers of factors. In the three factor
design, we use the model

$$y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k +
\left(\tau \beta\right)_{ij} + \left(\tau \gamma\right)_{ik} + \left(\beta
\gamma\right)_{jk} + \left(\tau \beta \gamma\right)_{ijk} + \epsilon_{ijkl}$$

where $\epsilon_{ijkl} \sim N\left(0, \sigma^2\right)$. Suppose that the first,
second, and third factors have $a, b$, and $c$ levels, respectively.

2. We’re dangerously close to getting lost in index purgatory, but notice certain
symmetries,
    * We have main effects for each factor
    	* $\tau_i, \beta_j, \gamma_k$
    * We have two-way interactions for each pair of factors
    	* $\left(\tau\beta\right)_{ij}, \dots$
    * We have a three-way interaction, between all factors
    	* $\left(\tau\beta\gamma\right)_{ijk}$

```{r, fig.margin = TRUE, echo = FALSE}
display(readImage("https://uwmadison.box.com/shared/static/b42kbwxdggtmf2krqr7uoj5r23fglfuo.png"))
display(readImage("https://uwmadison.box.com/shared/static/gnoa5i4w5qp6ehbweh30n3a2v91qpi89.png"))
display(readImage("https://uwmadison.box.com/shared/static/yqqhfc2kzk3egfvad8dvufed09t9m3tx.png"))
display(readImage("https://uwmadison.box.com/shared/static/goaewwcs61ewtrmfp4pcymzje1fdrbsb.png"))
```
    
3. We can calculate sum-of-squares terms for each of the terms. Notice that
there are also certain symmetries in the degrees of freedom,
    * $SS_A = a - 1$
    * $SS_B = b - 1$
    * $SS_C = c - 1$
    * $SS_{AB} = (a - 1)(b - 1)$
    * $SS_{BC} = (b - 1)(c - 1)$
    * ...
    * $SS_{ABC} = (a - 1)(b - 1)(c - 1)$
    
    What do you think is the pattern for arbitrary $K$.

4. For testing, we will compare these sums-of-squares to $SS_E$, which has
$abc(n - 1)$ degrees of freedom. The $F$-statistics for any of the terms above
can be found by dividing the associate mean squares against $MS_E$. Hence, we
can test whether any of the terms is nonzero for at least one value of its
index.

# Data Example

5. Let's look at a $2^3$ design (3 factors with two levels each). The goal is to
see how the etch rate on a chip varies as you change (A) gap between electrodes,
(B) power level, and (C) gas flow rate.

```{r}
plasma <- read.table("https://uwmadison.box.com/shared/static/f3sggiltyl5ycw1gu1vq7uv7omp4pjdg.txt", header=TRUE)
```

6. Looking at the data, there seems to be a strong interaction between A (the
x-axis) and C (the pairs of columns): the slope of the effect of A switches when
you go from one C configuration to the other.

```{r}
ggplot(plasma) +
  geom_point(aes(A, Rate)) +
  facet_grid(B ~ C)
```

7. We can quantify the strength of these relationships by estimating the model and
evaluating the relevant $F$-statistics. The `*` syntax refers to all main and
interaction effects derived from the linked variables.

```{r}
fit <- lm(Rate ~ A * B * C, plasma)
summary(aov(fit))
```
