---
title: "General Factorial Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [5.4](), [Rmarkdown]()_

```{r, echo = FALSE}
library(dplyr)
library(ggplot2)
library(EBImage)
library(readr)
theme_set(theme_bw())
```

1. We’ll discuss three factor factorial designs, with the hope that what we
learn will generalize to arbitrary numbers of factors. In the three factor
design, we use the model

$$y_{ijkl} = \mu + \tau_i + \beta_j + \gamma_k +
\left(\tau \beta\right)_{ij} + \left(\tau \gamma\right)_{ik} + \left(\beta
\gamma\right)_{jk} + \left(\tau \beta \gamma\right)_{ijk} + \epsilon_{ijkl}$$

where $\epsilon_{ijkl} \sim N\left(0, \sigma^2\right)$. Suppose that the first,
second, and third factors have $a, b$, and $c$ levels, respectively.

2. We’re dangerously close to getting lost in index purgatory, but notice certain
symmetries,

```{r, fig.margin = TRUE, echo = FALSE}
#display(readImage("https://uwmadison.box.com/shared/static/1mun8fg94m86cl9yxm1u7597js1jdcdg.png"), raster = TRUE)
#display(readImage("https://uwmadison.box.com/shared/static/xybpzb3td8ojijuidng5ghpbycea6lz8.png"), raster = TRUE)
#display(readImage("https://uwmadison.box.com/shared/static/8bs3ypma9dl3mh7amnlnxrxb0sgnmxho.png"), raster = TRUE)
#display(readImage("https://uwmadison.box.com/shared/static/t1jgtvnjvjw8vadc5swsu835zpenh6yi.png"), raster = TRUE)
```

    * We have main effects for each factor
    	* $\tau_i, \beta_j, \gamma_k$
    
    * We have two-way interactions for each pair of factors
    	* $\left(\tau\beta\right)_{ij}, \dots$
    	
    * We have a three-way interaction, between all factors
    	* $\left(\tau\beta\gamma\right)_{ijk}$
    
3. We can calculate sum-of-squares terms for each of the terms. Notice that
there are also certain symmetries in the degrees of freedom,
    * $SS_A = a - 1$
    * $SS_B = b - 1$
    * $SS_C = c - 1$
    * $SS_{AB} = (a - 1)(b - 1)$
    * $SS_{BC} = (b - 1)(c - 1)$
    * ...
    * $SS_{ABC} = (a - 1)(b - 1)(c - 1)$
    
    What do you think is the pattern for arbitrary $K$.

4. For testing, we will compare these sums-of-squares to $SS_E$, which has
$abc(n - 1)$ degrees of freedom. The $F$-statistics for any of the terms above
can be found by dividing the associate mean squares against $MS_E$. Hence, we
can test whether any of the terms is nonzero for at least one value of its
index.

# Data Example

5. Let's look at a $2^3$ design (3 factors with two levels each). The goal is to
see how the etch rate on a chip varies as you change (A) gap between electrodes,
(B) power level, and (C) gas flow rate.

```{r}
plasma <- read.table("https://uwmadison.box.com/shared/static/f3sggiltyl5ycw1gu1vq7uv7omp4pjdg.txt", header=TRUE)
```

6. Looking at the data, there seems to be a strong interaction between A (the
x-axis) and C (the pairs of columns): the slope of the effect of A switches when
you go from one C configuration to the other.

```{r}
ggplot(plasma) +
  geom_point( aes(x = A, y = Rate) ) +
  facet_grid(B ~ C)
```

7. We can quantify the strength of these relationships by estimating the model and
evaluating the relevant $F$-statistics. The `*` syntax refers to all main and
interaction effects derived from the linked variables.

```{r}
fit <- lm(Rate ~ A * B * C, plasma)
anova(fit)
```
