---
title: "Split-Plot Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-17-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [14.4](), [Rmarkdown]()_


```{r}
library("dplyr")
library("ggplot2")
library("readr")
library("EBImage")
library("reshape2")
theme_set(theme_bw() + theme(legend.position = "bottom"))
```
```{r, fig.cap = "A single field from a true factorial design in the irrigation example."}
#display(readImage("https://uwmadison.box.com/shared/static/qtxsckab5vcxffrhbpepj9peq28m7zwn.png"))
```

```{r, fig.cap = "An alternative split plot structure. The irrigation strategies are now rows, and each cell is a subplot."}
#display(readImage("https://uwmadison.box.com/shared/static/9s2cmkmgc45po62yzm8eltry8cy7rv42.png"))
```

There are times when we would like to perform a two-factor factorial experiment
across blocks, but one of the factors is much more difficult to vary than the
other. For example,

* We want to assess the effect of different irrigation ($A$) and corn varieties
($B$) on total yield. There are three fields (Blocks) within which to gather data.
It is hard to change irrigation strategy on small subplots of land, especially
compared to corn variety.

* We run a papermill, and want to compare pulp preparation strategies ($A$) and
cooking temperatures ($B$). Samples are collected over three days (Blocks). It is
hard to change the pulp preparation strategy from sample to sample — we would
rather prepare a few big batches —  but it’s easy to cook them at different
temperatures.

This makes a true $2^{2}$ factorial experiment impractical, because it would
require randomizing over all combinations of $A$ and $B$ for every sample that
we collect.

* If we had 3 irrigation strategies and 6 corn varieties, we would need to
divide each field into 18 subplots and randomize the assignment of irrigation x
corn pairs.

* It’s much easier to first divide each field into 3 large plots, and then
randomly assign corn varieties to 6 subplots within each large plot.

Effectively, practical considerations impose a restriction on randomization.

### Model

The model for a split-plot design is

\begin{align*}
y_{ijk} &= \mu + \tau_{k} + \alpha_{i} + \beta_{j} + \left(\tau\alpha\right)_{ki} + \left(\alpha\beta\right)_{ij} + \epsilon_{ijk}
\end{align*}

where $\epsilon \sim N\left(0, \sigma^2\right)$ independently.

The terms can be interpreted as,

* $\mu$: The global response average.

* $\tau_{k}$: The effect of the $k^{th}$ block (e.g., $k^{th}$ field).

* $\alpha_{i}$: The effect of the $i^{th}$ level of $A$, the
hard-to-granularly-randomize factor (e.g., $i^{th}$ irrigation strategy).

* $\beta_{j}$: The effect of the $j^{th}$ level of $B$, the
easy-to-granularly-randomize factor (e.g., $j^{th}$ corn variety)

* $\left(\tau\alpha\right)_{ki}$: An interaction factor between the $k^{th}$
block and the $i^{th}$ level of $A$ (e.g., the $i^{th}$ irrigation strategy
within the $k^{th}$ field might be unusually good).

* $\left(\alpha\beta\right)_{ij}$: An interaction factor between the $i^{th}$
level of $A$ and the $j^{th}$ level of $B$.

We will typically not care about individual block effects, though we will care
about the two different treatments. Therefore, it is common to

* Use random effects for $\tau_{k}, \left(\tau\alpha\right)_{ki}$.

* Use fixed-effects for $\alpha_{i}$ and $\beta_{j}$.

The expected mean squares associated with each of the terms above can be derived
in closed form. Here, we will simply illustrate their use through the `lme4`
package. The data are from the papermill experiment described above.

```{r}
pulp <- read_csv("https://uwmadison.box.com/shared/static/843r3mda46is46nbb5b6393u7caz8orq.csv") %>%
  mutate_at(vars(Day, Method, Temperature), as.factor)
```
```{r, fig.cap = "Data from the papermill experiment."}
ggplot(pulp) +
  geom_point(
    aes(x = Temperature, y = Strength, col = Day),
    size = 3
  ) +
  scale_color_brewer(palette = "Set2") +
  facet_grid(Method ~ .)
```

```{r}
fit <- aov(Strength ~ Method * Temperature + Error(Day / Method), data = pulp)
summary(fit)
```

### Split-Plot is not $2^{2}$ factorial

Fitting a $2^{2}$ factorial model when the data were collected with restrictions
on randomization can lead to misleading results. The code below fits an ordinary
$2^{2}$ factorial model to the papermill data. Note the overconfidence about an
effect of `Method`.

```{r}
fit <- aov(Strength ~ Method * Temperature, data = pulp)
summary(fit)
```