---
title: Following-up Two-Factor Fits
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library("knitr")
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [5.3](), [Rmarkdown]()_

```{r}
library(dplyr)
library(broom)
library(emmeans)
library(ggplot2)
library(readr)
theme_set(theme_bw())
```

1. We’ll cover the analogs of performing multiple comparisons, and doing model
diagnostics for the two-factor model. Almost everything will be familiar from
our experience with single-factor models.

```{r}
# background from last time
battery <- read_table2("https://uwmadison.box.com/shared/static/vmxs2wcsdxkdjujp85nw5kvk83xz4gl9.txt") %>%
  mutate(
    Material = as.factor(Material),
    Temperature = as.factor(Temperature)
  )
fit <- lm(Life ~ Material * Temperature, data=battery)
```

## Multiple Comparisons

2. We may want to use contrasts, to find out exactly how a particular factor is
associated with the response.

```{r, fig.margin=TRUE, echo = FALSE}
library("EBImage")
#display(readImage("https://uwmadison.box.com/shared/static/3j76s251jeervxrqheo8dqrm5vvq7e63.png"), method = "raster")
```

    * _Subtlety_: If the factor under investigation interacts with the other one,
    its effects will depend on that other factor.
    
    * _Solution_: Fix a level for the other factor, and study the influence of
    levels for the factor of interest.
    
    * Example: Fix temperature, and use Tukey’s HSD to study pairwise difference
    between materials, for that fixed temperature.

```{r}
emmeans(fit, pairwise ~ Material | Temperature)$contrasts
```

## Model Checking

3. Our key assumptions are independence, normality, and equal variances for the
$\epsilon_{ijk}$’s. Our diagnostics are based on residuals.

```{r, fig.margin = TRUE, echo = FALSE}
battery <- battery %>%
  mutate(
    resid = resid(fit),
    y_hat = predict(fit)
  )
ggplot(battery) +
  geom_point(aes(x = Temperature, y = resid))
```

  * Plot residuals against each of the two factors

```{r, fig.margin = TRUE, echo = FALSE}
ggplot(battery) +
  geom_point(aes(x = y_hat, y = resid))
```

  * Plot fitted vs. residual value

```{r, fig.height = 3, fig.width = 3.5, fig.margin = TRUE, echo = FALSE}
qqnorm(battery$resid, main = "")
qqline(battery$resid, col = "red")
```

  * Make normal probability plots
 
## Choosing the Sample Size

4. How should you choose how many replicates to have at each combination of the
two factors?

    * Simulate a model using a particular configuration of coefficients.
    * See how your power to detect effects varies as you increase the sample
    size.

```{r, echo = FALSE}
tau <- c(0, 1, 2)
beta <- c(0, 1)
ns <- seq(2, 10, by = 2)
b <- 1
sims <- list()
for (k in seq_along(ns)) {
  for (sim_rep in seq_len(50)) {
    for (i in seq_along(tau)) {
      for (j in seq_along(beta)) {
        sims[[b]] <- data.frame(
            "factor_1" = i,
            "factor_2" = j,
            "sample_size" = ns[k],
            "replicate" = seq_len(ns[k]),
            "value" = rnorm(ns[k], tau[i] + beta[j]),
            "sim_rep" = sim_rep
        )
        b <- b + 1
      }
    }
  }
}
sims <- bind_rows(sims)
```


```{r fig.margin = TRUE, fig.height = 5, fig.width = 2, echo = FALSE}
ggplot(sims %>% filter(sim_rep == 1)) +
  geom_point(aes(x = factor_1, y = value)) +
  facet_grid(sample_size ~ factor_2)
power <- sims %>%
  group_by(sample_size, sim_rep) %>%
  do(model = tidy(lm(value ~ factor_1 + factor_2, data = .))) %>%
  tidyr::unnest()
```


```{r echo = FALSE}
ggplot(power) +
  geom_point(
    aes(x = sample_size, y = p.value),
    position = position_jitter(w = .5),
    size = 0.5
  ) +
  geom_hline(yintercept = 0.05, col = "red") +
  facet_wrap(~ term)
```

```{r, fig.fullwidth = TRUE, echo = FALSE}
#display(readImage("https://uwmadison.box.com/shared/static/sie5e1b8v005l5ow8wzbtpehfdpn7wnf.png"), method = "raster")
```


## Interactions

5. If we only have one replicate per cell, then we can’t estimate an interaction
effect. If we tried, we’d be able to perfectly fit the data, so there would be
no way to estimate $\sigma^2$.

6. As a general recipe, to check for interactions, we can perform residual
analysis on the main effects model. Alternatively, we coudl use Tukey's
additivity test, which checks whether a multiplicative form of the interaction
is present, i.e., does 
$$y_{ijk} = \mu + \tau_i + \beta_j + \gamma \tau_i \beta_j$$ 
fit significantly better than the main effects model?
