---
title: "Projection and Blocking in $2^{K - p}$ Designs"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-17-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

_Readings [8.3, 8.4](), [Rmarkdown]()_


```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(EBImage)
library(reshape2)
library(ggplot2)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```
```{r, fig.cap = "Our earlier picture of projection from 3 to 2 factors in a complete factorial design."}
display(readImage("https://uwmadison.box.com/shared/static/b94hfeyugeun1xywlm2ccls9am5l38xg.jpg"))
```

To be fluent in working with fractional factorial designs, it’s worth knowing
how to project and block them. They provide useful alternative perspectives of
the designs we have already constructed.

### Projection

* Projection makes it possible to turn a fractional factorial screening
experiment into a full factorial experiment

* Effectively, you delete the columns of your design matrix involving the
factors outside the ones you want to project down to

* Depending on what design you project, a $2^{K - p}$ design will become
either (1) a full factorial design or (2) a $2^{K - r}$ fractional factorial
design, where $r < p$.

* Beware that conclusions will be somewhat tentative, because effects in the
full factorial could be aliased with high-order effects from factors that have
been projected out

Warm-up: (*Projecting a $2 ^ {4 - 1}$ Design*). Consider the $2 ^ {4 - 1}$
design with defining relation $I = ABCD$.

\newpage

| A | B | C | D |
| --- | --- | --- | --- |
| - | - | - | - |
| + | - | - | + |
| - | + | - | + |
| - | - | + | + |
| + | + | - | - |
| + | - | + | - | 
| - | + | + | - |
|+ | + | + | + |

What happens when you project out A?

| B | C | D |
| --- | --- | --- |
| - | - | - |
| - | - | + |
| + | - | + |
| - | + | + |
| + | - | - |
| - | + | - | 
| + | + | - |
| + | + | + |

You can check that this has all corners of the cube B-C-D, so it’s a full $2^{3}$ design.

Example (*Different projections of $2 ^ {6 - 2}$ design*). Let's consider the
example from our previous lecture

\newpage

| A | B | C	| D | E = ABC | F = BCD |
| --- | --- | --- | --- | --- | --- |
| - | - | - | - | - | - |
| + | - | - | - | + | - |
| - | + | - | - | + | + |
| - | - | + | - | + | + |
| - | - | - | + | - | + |
| + | + | - | - | - | + |
| + | - | + | - | - | + |
| + | - | - | + | + |  + 
| - | + | + | - | - | - |
| - | + | - | + | + | - |
| - | - | + | + | + | - |
| + | + | + | - | + | - |
| + | + | - | + | - | - |
| + | - | + | + | - | - |
| - | + | + | + | - | + |
| + | + | + | + | + | + |

* The study examined 6 factors over 16 runs. 

* The complete defining relations were $I = ABCE = BCDF = ADEF$.

* What happens when we project out $A$?

| B | C| D | E | F |
| --- | --- | --- | --- | --- | --- |
 | - | - | - | - | - |
 | - | - | - | + | - |
 | + | - | - | + | + |
 | - | + | - | + | + |
 | - | - | + | - | + |
 | + | - | - | - | + |
 | - | + | - | - | + |
 | - | - | + | + |  + |
 | + | + | - | - | - |
 | + | - | + | + | - |
 | - | + | + | + | - |
 | + | + | - | + | - |
 | + | - | + | - | - |
 | - | + | + | - | - |
 | + | + | + | - | + |
 | + | + | + | + | + |

Each configuration appears once. Since we have 16 rows still, this is evidently
a $2^{5 - 1}$ fractional factorial design.
 
What happens if we project out $E$ and $F$ instead? 

* This is exactly the original "basic design." I.e., the the full $2 ^ 4$
factorial before using the generators.

We could try to project out two factors that don't have such an obvious answer. So let's project out $A$ and $B$.

* It turns out we still get a $2 ^ 4$ full factorial, but now on $C$ through $F$.

| C| D | E | F |
| --- | --- | --- | --- | --- | --- |
 | - | - | - | - |
 | - | - | + | - |
 | - | - | + | + |
 | + | - | + | + |
 | - | + | - | + |
 | - | - | - | + |
 | + | - | - | + |
 | - | + | + |  + |
 | + | - | - | - |
 | - | + | + | - |
 | + | + | + | - |
 | + | - | + | - |
 | - | + | - | - |
 | + | + | - | - |
 | + | + | - | + |
 | + | + | + | + |
 
You might start suspecting that whenever we project out two variables, we will
end up with a full $2^{4}$ factorial. But this is not the case! 

* Consider projecting out $A$ and $E$.
* Now we have a $2 ^ {4 - 1}$ fractional factorial with 2 replicates. 
* Only half of the corners are present, but each one that appears is present
twice.

| B | C | D | F |
| --- | --- | --- | --- | --- | --- |
 | - | - | - | - |
 | - | - | - | - |
 | + | - | - | + |
 | - | + | - | + |
 | - | - | + | + |
 | + | - | - | + |
 | - | + | - | + |
 | - | - | + |  + |
 | + | + | - | - |
 | + | - | + | - |
 | - | + | + | - |
 | + | + | - | - |
 | + | - | + | - |
 | - | + | + | - |
 | + | + | + | + |
 | + | + | + | + |
 
This ends up being a general rule. If none of the defining words contain all the
factors that you’re projecting out, then you’re left with a full factorial
design; otherwise you get a fractional factorial with replicates.

* E.g., $AE$ is part of the first defining relation, so we got a fractional
factorial.

Exercise: Try projecting onto $A, B, C$. Is it a fractional or full factorial? How
many replicates each? Could you have determined this, just by looking at the
defining relations?

* A design is said to have _projectivity_ $v$ is collapsing to any subset of
$v$ factors will result in a full factorial design. The example above has
projectivity 3.

### Blocking

Let’s say we’ve subsetted to a set of runs corresponding to a fractional
factorial design. But, we still have to break the runs into batches, we have too
many runs to be able to reasonably run them in one block. How can we
appropriately block a fractional factorial design?

* The typical solution is to choose a high-order effect and alias it with the
batches. This divides the runs into two blocks, those where the effect is + and
those where it is -.

* Ensure that the alias group doesn’t contain any low-order effects of interest
(e.g., we don’t want the blocks to be confounded with any main effects)

* We effectively sacrifice the alias group that is confounded with the block.

Example: For our $2 ^{6 - 2}$ design above, we can confound $ACD$ with the
block, to avoid aliasing any main effects with the block effect. This let us
divide our runs into two blocks of size 8.

| A | B | C	| D | E | F | Block = ACD |
| --- | --- | --- | --- | --- | --- | --- |
| - | - | - | - | - | - | - (B1) | 
| + | - | - | - | + | - | + (B2) |
| - | + | - | - | + | + | - (B1) |
| - | - | + | - | + | + | + (B2) |
| - | - | - | + | - | + | + (B2) |
| + | + | - | - | - | + | + (B2) |
| + | - | + | - | - | + | - (B1) |
| + | - | - | + | + |  + | - (B1) |
| - | + | + | - | - | - | + (B2) |
| - | + | - | + | + | - | + (B2) |
| - | - | + | + | + | - | - (B1) |
| + | + | + | - | + | - | - (B1) |
| + | + | - | + | - | - | - (B1) |
| + | - | + | + | - | - | + (B2) |
| - | + | + | + | - | + | - (B1) |
| + | + | + | + | + | + | + (B2)|