---
title: "Mixture Experiments"
description: |
  Optimizing responses when mixing ingredients.
author:
  - name: Kris Sankaran
    url: {}
date: 12-09-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = FALSE)
```

_Readings [11.6](https://www.wiley.com/en-us/Design+and+Analysis+of+Experiments%2C+10th+Edition-p-9781119492443#content-section), [Rmarkdown](https://github.com/krisrs1128/stat424_f21/blob/main/_posts/2021-08-17-week13-3/week13-3.Rmd)_


```{r}
library(ggplot2)
theme424 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(theme424)
```

1. There are many situations where we may want to find a mixture of
ingredients to optimize some criteria. For example, when making a fiber for a
kind of yarn, there are three ingredients (polyethylene, polystyrene, and
polypropylene) that are mixed at various fractions. What fraction would optimize
the stretchiness of the yarn?

2. We’ll explore this problem as a special case of the general response surface
problem.

### Geometry

3. The mixture setting induces specific geometric constraints. Suppose there are
$P$ total ingredients, and let $x_{p}$ denote the fraction of ingredient $p$.
Since the $x_{p}$ are mixture fractions, we have the constraints,
  * $x_{p} \in \left[0, 1\right]$
  * $\sum_{p = 1}^{P} x_{p} = 1$
  
4. The set of $x = \left(x_{1}, \dots, x_{P}\right)$ that satisfy these constraints
can be geometrically represented by a simplex.
  * Center point has an equal amount of each ingredient
  * Corners have 100% coming from one of the ingredients
  There is nothing stopping us from fitting a response surface over the simplex.

### Design Points

4. What design points should we use? Factorial and central composite designs
don't respect the geometry of the simplex, so we need alternatives.

```{r, fig.cap = "Examples of simplex lattice and centroid designs." }
library(mixexp)
DesignPoints(SLD(3, 3))
DesignPoints(SLD(3, 5))
DesignPoints(SCD(3))
```

5. Simplex lattice design: Choose some $m$ which will reflect the granularity of
our design. Consider combinations of integers $k_{p} \in \{0,1, \dots, m\}$ such
that $\sum_{p= 1}^{P} k_{p} = m$. Each such combination specifies a point
\begin{align*}
\frac{1}{m}\left(k_{1}, \dots, k_{P}\right)
\end{align*}
	that is included in the simplex lattice design. An alternative way to specify
	this design is to draw a lattice grid with spacing $\frac{1}{m}$ in the unit
	cube, and then discard points that don’t lie in the simplex.
	
6. Simplex centroid design
	* Corners: Add all $P$ permutations of the vector $\left(1, 0, \dots, 0\right)$.
	* Edge midpoints: Add all ${P \choose 2}$ permutations of $\left(\frac{1}{2}, \frac{1}{2}, 0, \dots, 0\right)$. These are midpoints between two corners, and so lie on edges of the simplex.
	* Face centroids: Add all ${P \choose 3}$ permutations of $\left(\frac{1}{3}, \frac{1}{3}, \frac{1}{3}, 0, \dots, 0\right)$ which are the centers of faces defined by three corners.
	* .. continue the pattern: For all $k \leq P$, add all ${P \choose k}$ permutations of $\left(\frac{1}{k}, \dots, \frac{1}{k}, 0, \dots, 0\right)$.
	
7. The block below shows the first few rows from both a simplex centroid (`SCD`)
and simplex linear (`SLD`) design.

```{r}
options(echo = TRUE)
```

```{r}
library(mixexp)
head(SCD(5))
head(SLD(5, 2))
```
	
8. There are some common variations,
  * It’s common to augment the designs above with center points.
  * Sometimes it is useful to include _axial points_, which are samples along rays   extending from corners of the simplex
  * Computer-generated designs can be used, especially when there are constraints on feasible mixture values.

```{r, fig.cap = "Simplex centroid design across 5 mixture elements.", echo = FALSE}
mscd <- SCD(5) %>%
  mutate(id = row_number()) %>%
  pivot_longer(-id)

ggplot(mscd) +
  geom_tile(aes(name, id, fill = value)) +
  scale_fill_viridis_c() +
  scale_y_continuous(expand = c(0, 0)) +
  coord_fixed() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 8),
    axis.title = element_blank()
  )
```

### Data Example

9. We'll use the yarn data from Example 11.5. The experiment used a (3, 2)
simplex lattice design to measure variation in yarn elongation as a function of
the fractions of different materials used to make the base fiber. First, let's
try to see the dependence visually, though direct visualization on the simplex
is challenging.

```{r}
library(readr)
yarn <- read_csv("https://uwmadison.box.com/shared/static/jghwbsnn6qjpwdr1lc97p9mbxk8qkwif.csv")
ggplot(yarn) +
  geom_point(
    aes(x1, x2, size = x3, col = elongation),
    position = position_jitter(w = 0.1, h = 0.1)
  ) +
  scale_size(range = c(2, 5)) +
  scale_color_viridis_c() +
  theme(legend.position = "none")
```

10. Now, let's fit a second-order polynomial to the data. Note that we include a
`-1` term in the fit below, to ensure the model does not fit an intercept term.

```{r, size = "scriptsize"}
fit <- lm(elongation ~ -1 + (x1 + x2 + x3) ^ 2, data = yarn)
summary(fit)
```
We can plot the associated fit. Compare with Figure 11.43.

```{r, echo = TRUE, fig.margin = FALSE, fig.width = 10}
ModelPlot(
  fit, 
  dimensions = list(x1 = "x1", x2 = "x2", x3 = "x3"), 
  contour = TRUE
)
```
