---
title: "$2 ^ 3$ Factorial Design"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r, fig.margin = TRUE}
library(EBImage)
display(readImage("https://uwmadison.box.com/shared/static/sie5e1b8v005l5ow8wzbtpehfdpn7wnf.png"))
```

```{r}
library(dplyr)
library(ggplot2)
library(readr)
theme_set(theme_bw())
```

* How can we generalize the $2^2$ analysis so that we can do studies that
inspect many factors simultaneously (all at two levels each)?

* We’ll analyze the 3 factor case (the $2^3$ design) first, but in search of
general patterns

* For the $2^3$ design, we have 8 factor configurations

	* Visualize as corners of a cube
	
	* Call the third factor $C$.
	
```{r}
plasma <- read.table("https://uwmadison.box.com/shared/static/f3sggiltyl5ycw1gu1vq7uv7omp4pjdg.txt", header = TRUE) %>%
  mutate(
    A = as.factor(A),
    B = as.factor(B),
    C = as.factor(C)
  )
```

```{r, fig.height = 5, fig.margin = TRUE}
ggplot(plasma) +
  geom_point(
    aes(x = A, y = B, col = Rate),
    position = position_jitter(w = 0.1, h = 0.1)
  ) +
  facet_grid(C ~ .) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom")
```

## Effect Estimates

Our table notation can be extended to deal with all 8 corners of the cube.

| A | B | C | label |
|---|---|--- | ---------|
| - | - | - | (1) |
| + | - | - | a       |
| - | + | - | b       |
| - | - | + | c       |
| + | + | - | ab      |
| + | - | + | ac      |
| - | + | + | bc      |
| + | + | + | abc   |

* The main effect estimates can be made by subtracting the + from the - corners
 	* Average of when factor is on vs. off
	* Average of one face of the cube minus the average of the opposite

For example,
$$A = \frac{1}{2^2 n}\left[\left(a + ab + ac + abc\right) - \left(\left(1\right) + b + c + bc\right)\right]$$

To estimate interactions, look at how average effects change when you condition.

| B | Average A Effect | 
| --- | ------| 
| + | $\frac{1}{2^2 n}\left[\left(abc - bc\right) + \left(ab - b\right)\right]$ |
| - | $\frac{1}{2^2 n}\left[\left(ac - c\right) + \left(a - \left(1\right)\right)\right]$ |

Which inspires the definition,

$$
AB = \frac{1}{2^2 n}\left[abc - bc + ab - b - ac + c - a + \left(1\right)\right]
$$

And notice that the associated contrast can be obtained by multiplying the
columns in the table above.

```{r, fig.margin = TRUE}
display(readImage("https://uwmadison.box.com/shared/static/3j76s251jeervxrqheo8dqrm5vvq7e63.png"))
display(readImage("https://uwmadison.box.com/shared/static/8h0vt5kcb5ss0l9dlghsswnpf88q8267.png"))
```


| A | B | C | AB | label |
| --- | --- | --- | --- | ------| 
| - | - | - | + | (1) |
| + | - | - | - | a |
| - | + | - | - | b |
| - | - | + | + | c |
| + | + | - | + | ab |
| + | - | + | - | ac |
| - | + | + | - |  bc |
| + | + | + | + | abc |

We won’t prove why this works, but you can use it as a device for avoiding
having to memorize everything. The three-way interaction is defined as the
change in two-way interactions across the two values for the third variable.
It's contrast can be derived also by multiplying the relevant columns from the
table above.
