---
title: "$2 ^ 3$ Factorial Design"
description: |
  A short description of the post.
author:
  - name: Kris Sankaran
    url: {}
date: 08-16-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE, echo = TRUE)
```

```{r, fig.margin = TRUE}
library(EBImage)
display(readImage("https://uwmadison.box.com/shared/static/vxvzawk89r2m08um53dbfoka5md0bsnf.png"))
```

```{r}
library(dplyr)
library(ggplot2)
library(readr)
theme_set(theme_bw())
```

1. How can we generalize the $2^2$ analysis so that we can do studies that
inspect many factors simultaneously (all at two levels each)? For now, we’ll
analyze the 3 factor case (the $2^3$ design), but with an eye out for more
general patterns

2. For the $2^3$ design, we have 8 factor configurations
  	* Visualize as corners of a cube
  	* Call the third factor $C$.
	
```{r}
plasma <- read.table("https://uwmadison.box.com/shared/static/f3sggiltyl5ycw1gu1vq7uv7omp4pjdg.txt", header = TRUE) %>%
  mutate(
    A = as.factor(A),
    B = as.factor(B),
    C = as.factor(C)
  )
```

```{r, fig.height = 5, fig.margin = TRUE}
ggplot(plasma) +
  geom_point(
    aes(x = A, y = B, col = Rate),
    position = position_jitter(w = 0.1, h = 0.1)
  ) +
  facet_grid(C ~ .) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom")
```

### Effect Estimates

3. Our table notation can be extended to deal with all 8 corners of the cube.

| A | B | C | label |
|---|---|--- | ---------|
| - | - | - | (1) |
| + | - | - | a       |
| - | + | - | b       |
| - | - | + | c       |
| + | + | - | ab      |
| + | - | + | ac      |
| - | + | + | bc      |
| + | + | + | abc   |

4. The main effect estimates can be made by subtracting the + from the -
corners. Equivalently, this can be viewed as the difference in averages,
 	* when the factor is on vs. off
	* between one face of the cube and its opposite

5. For example, to estimate the main effect of $A$, we can use,
$$A = \frac{1}{2^2 n}\left[\left(a + ab + ac + abc\right) - \left(\left(1\right) + b + c + bc\right)\right]$$

6. To estimate interactions, we compare how the average effects of a variable
change when we condition on the value of another variable. For example, for the
interaction $AB$, notice that

| B | Average A Effect | 
| --- | ------| 
| + | $\frac{1}{2^2 n}\left[\left(abc - bc\right) + \left(ab - b\right)\right]$ |
| - | $\frac{1}{2^2 n}\left[\left(ac - c\right) + \left(a - \left(1\right)\right)\right]$ |

    which inspires the definition,

$$
AB = \frac{1}{2^2 n}\left[abc - bc + ab - b - ac + c - a + \left(1\right)\right]
$$

7. Notice that the associated contrast can be obtained by multiplying the
columns in the table above.

```{r, fig.margin = TRUE}
display(readImage("https://uwmadison.box.com/shared/static/wwsniua1ce0q376oy2ffrtzf4ivzup99.png"))
display(readImage("https://uwmadison.box.com/shared/static/hapjw92oqij5oiyddt29nvtq6y1xzxxu.png"))
```

  
  | A | B | C | AB | label |
  | --- | --- | --- | --- | ------| 
  | - | - | - | + | (1) |
  | + | - | - | - | a |
  | - | + | - | - | b |
  | - | - | + | + | c |
  | + | + | - | + | ab |
  | + | - | + | - | ac |
  | - | + | + | - |  bc |
  | + | + | + | + | abc |
    
  We won’t prove why this works, but you can use it as a device for avoiding
  having to memorize everything. The three-way interaction is defined as the
  change in two-way interactions across the two values for the third variable.
  It's contrast can be derived also by multiplying the relevant columns from the
  table above.
  