---
title: "$2^K$ Design Case Studies"
author: "Kris Sankaran | UW Madison | 23 September 2021"
output:
  xaringan::moon_reader:
    css: ["default", "css/xaringan-themer.css", "cols.css"]
    lib_dir: libs
    nature:
      beforeInit: "cols_macro.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
    seal: false  
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, dpi = 200, fig.width = 5, fig.height = 3, dev = 'svg', dev.args = list(bg = "transparent"))
theme424 <- theme_minimal() + 
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", size = 0.6),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.position = "bottom"
  )
theme_set(theme424)
```

# $2^K$ Design Case Studies

```{r, out.width = 200}
```
### Statistical Experimental Design

.large[Kris Sankaran | UW Madison]

---

### Today

* Book Sections: 6.6
* Online Notes: Week 9 [2]

---

### Motivation

* Analysis of factorial experiments is hard to completely automate
* Paying attention to structures in the problem or data can pay dividends
* We'll consider three case studies to illustrate typical scenarios
* We show code along the way

---

# Example 6.3

---

### Drill Dataset

An experiment was done to see how the "advance rate of a drill" varied as a
function of four factors, which we will call A, B, C, and D.

```{r}
library(dplyr)
library(readr)
code <- function(x) ifelse(x == '-', -1, 1)
drill <- read_csv("https://uwmadison.box.com/shared/static/7l8bpcu36a12a8c0chlh4id0qezdnnoe.csv") %>%
  mutate_at(vars(A:D), code)
head(drill, 4)
```

---

### Daniel Plot

.pull-left[
* We first fit a full $2^4$ model and get effects by doubling the regression
coefficients.
* It seems like we should drop factor A
]

```{r, echo = FALSE}
daniel_plot <- function(effects, probs = c(0.3, 0.7)) { 
  qq <- qqnorm(effects, datax = TRUE)
  qqline(effects, col = "red", probs = probs, datax = TRUE)
  text(qq$x, qq$y, names(effects), pos=1)
}
```

.pull-right[
```{r, fig.height = 5, fig.width = 5}
fit <- lm(rate ~ A * B * C * D, drill)
effects <- 2 * coef(fit)[-1]
daniel_plot(effects, c(0.35, 0.65)) # same code from last lecture
```
]

---

### Fitted Submodel

* We fit the submodel without A: `rate ~ B * C * D`.
* We use `resid` and `predict` to get residual and predicted values from the
submodel

```{r}
fit <- lm(rate ~ B * C * D, drill) # removed A from the model
drill_resid <- drill %>%
  mutate(residual = resid(fit), y_hat = predict(fit))
```

---

### Submodel Residuals

* Larger residuals are associated with higher predicted values.
* We should transform the original rates.
  - Let's try `log`, which is often useful for nonnegative or skewed data

```{r}
ggplot(drill_resid) +
  geom_point(aes(y_hat, residual))
```

---

### New Daniel Plot

.pull-left[
* We can regenerate the Daniel plot using the transformed response.
* B, C, and D alone now seem to strongly stand out
  - We should fit the submodel with just these factors
]

.pull-right[
```{r, fig.height = 5, fig.width = 5}
fit <- lm(log(rate) ~ A * B * C * D, data = drill)
daniel_plot(2 * coef(fit)[-1])
```
]

---

### New Submodel

The residuals of the new submodel are much better behaved.

```{r}
fit <- lm(log(rate) ~ B + C + D, drill)
drill_resid <- drill %>%
  mutate(residual = resid(fit), y_hat = predict(fit))
ggplot(drill_resid) +
  geom_point(aes(y_hat, residual))
```

---

### New Submodel

The ANOVA table quantifies the significance of the effects we drew out by eye.
They also guide us to a plot where the effects are clearer.

.pull-left[
```{r}
summary(aov(fit))
```
]

.pull-right[
```{r, fig.height = 2.5}
ggplot(drill) +
  geom_point(aes(B, log(rate), col = as.factor(D))) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~ C)
```
]

---

# Example 6.4

---

### Airplane Window Dataset

How do temperature (A), clamp time (B), resin flow (C), and press closing time
(D) affect airplane window defect rates?

```{r}
code <- function(x) ifelse(x == '-', -1, 1)
windows <- read_csv("https://uwmadison.box.com/shared/static/62phufkeprheu9gu35mu1e75x6rc2shv.csv") %>%
  mutate_at(vars(A:D), code)
```

---

### Effect Estimates

.pull-left[
We make a Daniel plot from the full data. The story seems simple,
  * Temperature (A) has a strong positive effect
  * Resin (C) flow has a slight negative effect.
]

.pull-right[
```{r, fig.height = 5, fig.width = 5}
fit <- lm(defects ~ A * B * C * D, data = windows)
daniel_plot(2 * coef(fit)[-1])
```
]
  
---

### Fitted Submodel

The A effect is very strong, and both A and C are significant.

```{r}
fit <- lm(defects ~ A + C, data = windows)
summary(aov(fit))
```

---

### Interpretation

.pull-left[
Both the coefficients and the plot suggest that turning A off, $A = -1$, and C
on, $C = 1$, will minimize the defect rate.
]

.pull-right[
```{r}
coef(fit)
ggplot(windows) +
  geom_point(aes(A, defects, col = as.factor(C))) +
  scale_color_brewer(palette = "Set2")
```
]

---

### Residual Plot

Plotting residuals of the submodel against factors, we notice strong heteroskedasticity.
* B does not affect the mean, but does affect the error variance.

```{r, fig.height = 2.5}
windows$residual <- resid(fit)
ggplot(windows) +
  geom_point(aes(B, residual))
```

---

### Approach

Make the recommendation A off, C on, B off

B has no effect on the average defect rate, but keeping it off will reduce the
variance in defect rate around the optimal A and C values.

Takeaway: In practice, it’s often useful to take variability into account,
rather than just average response.

---

# Example 6.5

---

### Semiconductor Dataset

How do temperature (A), time (B), pressure (C), and gas flow (D) affect oxide
thickness of semiconductor wafers? The code below downloads the data and
reshapes it so that the oxide thickness variable is all in one column.

```{r}
library(tidyr)
oxide <- read_csv("https://uwmadison.box.com/shared/static/vyk6uoe3zbnonv4n6jcusbrocmt4cvru.csv") %>%
  pivot_longer(starts_with("wafer"), names_to = "variable")
head(oxide, 4)
```

---

### Repeated Measures

* In the design, four wafers are put in the furnace at a time.
* These are not independent replicates, they are repeated measures.
* We average the responses within each factor combination
  - Pretend it is an unreplicated design

```{r}
oxide_collapse <- oxide %>%
  group_by(A, B, C, D) %>% # isolate independent configurations
  summarise(mean = mean(value), var = var(value)) # take average and var. across groups
head(oxide_collapse, 4)
```

---

### Daniel Plot

.pull-left[
A, B, C, and the interactions AB and AC are likely nonnull.
]

.pull-right[
```{r, fig.width = 5, fig.height = 5}
fit <- lm(mean ~ A * B * C * D, data = oxide_collapse)
daniel_plot(2 * coef(fit)[-1])
```
]

---

### Submodel

.pull-left[
To fit the all the main effects together when the interactions AB, AC, we can
use the formula `~ A * (B + C)`.
]
.pull-right[
```{r}
fit <- lm(mean ~ A * (B + C), oxide_collapse)
summary(fit) # the p-values are the same as for summary(aov(fit))
```
]

---

### Caution

.pull-left[
* If we had imagined that the repeated measures were actually replicates, we
would have identified many false positives.
* From the Daniel plot, factor D is almost definitely not significant
]

.pull-right[
```{r}
fit <- lm(value ~ A * B * C * D, data = oxide)
summary(fit)
```
]

---

### Visualization

The changes in slope across colors and panels are exactly the AB and AC
interaction effects.

```{r}
ggplot(oxide_collapse) +
  geom_point(aes(A, mean, col = as.factor(C))) +
  facet_wrap(~ B)
```

---

### Visualization: Response Surface

* We can also use `image` (in `rsm`) to view the fitted response
* This can be read like a topographic map -- darker colors are higher
* Here, the interaction appears as curvature

```{r}
library(rsm)
fit <- lm(mean ~ A * (B + C), oxide_collapse)
image(fit, ~ A + B + C)
```

---

### Modeling the Variances

Above, we modeled the average of the repeated measures within each factor
combination. We can instead model the variances.

The choice `~ A + B * D` came from looking at the Daniel plot.

```{r}
fit <- lm(var ~ A + B * D, data = oxide_collapse)
image(fit, ~ A + B + D)
```
